---
title: 'brms exercises: historical controls'
author: 'your-name-here'
date: today
output: 
  html_document:
    toc: true
    embed-resources: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Background

These exercises accompany the case study [Use of Historical Control Data](https://opensource.nibr.com/bamdd/src/02a_meta_analysis.html) as part of the training course "Workshop on applied modeling in drug development using brms."

## Preliminary code

First load the required packages and set some options as described in the case study.

```{r, warning = FALSE, message = FALSE}
here::i_am("exercises/ex1_historical_control.qmd")
library(here)

library(ggplot2)
library(dplyr)
library(knitr)
library(brms)
library(posterior)
library(bayesplot)
library(RBesT)
theme_set(theme_bw())

options(
  # how many processor cores would you like to use?
  mc.cores = 4, 
  # how would you like to access Stan?
  brms.backend = "cmdstanr",
  # cache model binaries
  cmdstanr_write_stan_file_dir=here::here("_brms-cache"),
  # no need to normalize likelihoods
  brms.normalize = FALSE,
  # when you are storing your model to file, 
  # how shall it be updated?
  brms.file_refit = "on_change"  
  # alternatives: "never", "always"
  # use "never" for production
)

set.seed(254335)
```

Then create the dataset used in the example:

```{r}
AS_region <- bind_cols(RBesT::AS, region=sample(c("asia", "europe", "north_america"), 8, TRUE))
kable(AS_region)
```

And fit the model as shown in the case study:

```{r}
model <- bf(r | trials(n) ~ 1 + (1 | study), family=binomial)

model_prior <- prior(normal(0, 2), class="Intercept") +
    prior(normal(0, 1), class="sd", coef="Intercept", group="study")

map_mc_brms  <- brm(model, AS_region, prior = model_prior,
                    seed = 4767,
                    silent = 2, refresh = 0)

```

## Exercise 1

Create a posterior predictive check based on the predictive
distribution for the response rate.  
Steps:
* Use `posterior_predict` to create samples from the predictive
  distribution of outcomes per trial.
* Use `sweep(predictive, 2, AS_region$n, "/")` to convert these
  samples from the predictive distribution of the outcome counts to
  samples from the predictive distribution for responder rates.
* Use `ppc_intervals` from `bayesplot` to create a plot showing your
  results.
    
```{r}
# Your solution here -----------------------------------------------------------


```


## Exercise 2

Redo the analysis with region, but treat region as a fixed
   effect. Evaluate the informativeness of the obtained MAP priors.
   The model formula for `brms` should look like `region_model_fixed
   <- bf(r | trials(n) ~ 1 + region + (1 | study), family=binomial)`.
Steps:
* Consider the prior for the region fixed effect first. The reference
  region is included in the intercept. The reference region is
  implicitly defined by the first level of the variable region when
  defined as `factor`.
  - Define `asia` to be the reference region in the example. Also
    include a level `other` in the set of levels.
  - Assume that an odds-ratio of $2$ between regions can be seen as
    very large such that a prior of $\mbox{N}(0, (\log(2)/1.96)^2)$
    for the region main effect is adequate.
* Obtain the MAP prior for each region by using the
  `AS_region_all` data frame defined below and apply
  `posterior_linpred` as shown above.
* Convert the MCMC samples from the MAP prior distribution into
  mixture distributions with the same code as above.
* Calculate the ESS for each prior distribution with the `ess`
  function from `RBesT`.
	
```{r}

AS_region_all <- data.frame(region=c("asia", "europe", "north_america", "other")) %>%
    mutate(study=paste("new_study", region, sep="_"), r=0, n=6)

# Your solution here -----------------------------------------------------------


```
	
   
## Exercise 3

Run the analysis for the normal endpoint in the `crohn` data set of
`RBesT`. Refer to the `RBesT` vignette for a [normal endpoint](https://cran.r-project.org/web/packages/RBesT/vignettes/introduction.html)
   on more details and context.  
Steps:
* Use as `family=gaussian` and use the `se` response modifier in place
  of `trials` to specify a known standard error.
* Use the same priors as proposed in the vignette.
* Compare the obtained MAP prior (in MCMC sample form) from `RBesT`
  and `brms`.
    
```{r}
crohn <- RBesT::crohn

# Your solution here -----------------------------------------------------------


```



