---
title: 'brms exercises: dose finding'
author: 'Björn Holzhauer'
date: today
output: 
  html_document:
    toc: true
    embed-resources: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Background

This exercises accompany the case study [Dose Finding](https://opensource.nibr.com/bamdd/src/02b_dose_finding.html) as part of the training course "Workshop on applied modeling in drug development using brms."

## Preliminary code

First load the required packages and set some options as described in the case study.

```{r, warning = FALSE, message = FALSE}
here::i_am("exercises/solutions/ex2_dose_finding.qmd")
library(here)
library(tidyverse)
library(brms)
library(tidybayes)
library(ggrepel)
library(patchwork)
library(ggplot2)
library(ggdist)

# Set defaults for ggplot2 ----
theme_set(theme_bw(base_size=18) +
          theme(legend.position = "bottom"))

scale_colour_discrete <- function(...) {
  # Alternative: ggsci::scale_color_nejm(...)
  # scale_colour_brewer(..., palette="Set1")
  ggthemes::scale_colour_colorblind(...)
}
scale_fill_discrete <- function(...) {
  # Alternative: ggsci::scale_fill_nejm(...)
  #scale_fill_brewer(..., palette="Set1")
  ggthemes::scale_fill_colorblind(...)
}
scale_colour_continuous <- function(...) {
  scale_colour_viridis_c(..., option="turbo")
}
update_geom_defaults("point", list(size=2))
update_geom_defaults("line", list(size=1.5))

options(
  # how many processor cores would you like to use?
  mc.cores = 4, 
  # how would you like to access Stan?
  brms.backend = "cmdstanr",
  # cache model binaries
  cmdstanr_write_stan_file_dir=here::here("_brms-cache"),
  # no need to normalize likelihoods
  brms.normalize = FALSE,
  # when you are storing your model to file, 
  # how shall it be updated?
  brms.file_refit = "on_change"  
  # alternatives: "never", "always"
  # use "never" for production
)

set.seed(254335)
```

## Excercise: 

### Exercise 1: Peanut allergy

In a hypothetical phase 2b trial children and adolescents with peanut allergy 
were randomly assigned to placebo or 5 doses (5 to 300 mg) of a new drug.  
After 26 weeks the patients underwent a double-blind placebo-controlled
food challenge and the primary endpoint was the proportion of patients that
could ingest 600 mg or more of peanut protein without experiencing dose-limiting
symptoms. The plots below show an overview of the data. Note that no placebo
patients were considered "responders" per the primary endpoint definition.

```{r peanutdata, class.source = 'fold-hide'}
peanut <- tibble(
  TRT01P = structure(c(
    6L, 6L, 6L, 6L, 6L, 6L, 6L, 6L, 6L, 6L, 6L, 6L, 6L, 6L, 6L, 6L, 6L, 6L, 6L, 6L, 6L, 6L, 6L, 6L, 6L, 6L, 6L, 6L,
    6L, 6L, 6L, 6L, 6L, 6L, 6L, 6L, 6L, 6L, 6L, 6L, 6L, 6L, 6L, 6L, 6L, 6L, 6L, 6L, 6L, 6L, 5L, 5L, 5L, 5L, 5L, 5L,
    5L, 5L, 5L, 5L, 5L, 5L, 5L, 5L, 5L, 5L, 5L, 5L, 5L, 5L, 5L, 5L, 5L, 5L, 5L, 5L, 5L, 5L, 5L, 5L, 5L, 5L, 5L, 5L,
    5L, 5L, 5L, 5L, 5L, 5L, 5L, 5L, 5L, 5L, 5L, 5L, 5L, 5L, 5L, 5L, 4L, 4L, 4L, 4L, 4L, 4L, 4L, 4L, 4L, 4L, 4L, 4L,
    4L, 4L, 4L, 4L, 4L, 4L, 4L, 4L, 4L, 4L, 4L, 4L, 4L, 4L, 4L, 4L, 4L, 4L, 4L, 4L, 4L, 4L, 4L, 4L, 4L, 4L, 4L, 4L,
    4L, 4L, 4L, 4L, 4L, 4L, 4L, 4L, 4L, 4L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L,
    3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L,
    3L, 3L, 3L, 3L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L,
    2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 1L, 1L,
    1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L,
    1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L
  ),levels = c("PBO", "5 mg", "15 mg", "50 mg", "150 mg", "300 mg"), class = "factor"),
  dose = c(
    300L, 300L, 300L, 300L, 300L, 300L, 300L, 300L, 300L, 300L, 300L, 300L, 300L, 300L, 300L, 300L, 300L, 300L, 300L,
    300L, 300L, 300L, 300L, 300L, 300L, 300L, 300L, 300L, 300L, 300L, 300L, 300L, 300L, 300L, 300L, 300L, 300L, 300L,
    300L, 300L, 300L, 300L, 300L, 300L, 300L, 300L, 300L, 300L, 300L, 300L, 150L, 150L, 150L, 150L, 150L, 150L, 150L,
    150L, 150L, 150L, 150L, 150L, 150L, 150L, 150L, 150L, 150L, 150L, 150L, 150L, 150L, 150L, 150L, 150L, 150L, 150L,
    150L, 150L, 150L, 150L, 150L, 150L, 150L, 150L, 150L, 150L, 150L, 150L, 150L, 150L, 150L, 150L, 150L, 150L, 150L,
    150L, 150L, 150L, 150L, 150L, 50L, 50L, 50L, 50L, 50L, 50L, 50L, 50L, 50L, 50L, 50L, 50L, 50L, 50L, 50L, 50L, 50L,
    50L, 50L, 50L, 50L, 50L, 50L, 50L, 50L, 50L, 50L, 50L, 50L, 50L, 50L, 50L, 50L, 50L, 50L, 50L, 50L, 50L, 50L, 50L,
    50L, 50L, 50L, 50L, 50L, 50L, 50L, 50L, 50L, 50L, 15L, 15L, 15L, 15L, 15L, 15L, 15L, 15L, 15L, 15L, 15L, 15L, 15L,
    15L, 15L, 15L, 15L, 15L, 15L, 15L, 15L, 15L, 15L, 15L, 15L, 15L, 15L, 15L, 15L, 15L, 15L, 15L, 15L, 15L, 15L, 15L,
    15L, 15L, 15L, 15L, 15L, 15L, 15L, 15L, 15L, 15L, 15L, 15L, 15L, 15L, 5L, 5L, 5L, 5L, 5L, 5L, 5L, 5L, 5L, 5L, 5L,
    5L, 5L, 5L, 5L, 5L, 5L, 5L, 5L, 5L, 5L, 5L, 5L, 5L, 5L, 5L, 5L, 5L, 5L, 5L, 5L, 5L, 5L, 5L, 5L, 5L, 5L, 5L, 5L, 5L,
    5L, 5L, 5L, 5L, 5L, 5L, 5L, 5L, 5L, 5L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L,
    0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L,
    0L, 0L),
  AVAL = c(
    1000L, 100L, 0L, 600L, 600L, 300L, 300L, 10L, 10L, 100L, 1000L, 300L, 10L, 1000L, 1000L, 1000L, 3L, 3L, 600L, 300L,
    1000L, 1000L, 100L, 1000L, 600L, 600L, 1000L, 1000L, 1000L, 300L, 600L, 1000L, 1000L, 1000L, 1000L, 1000L, 300L,
    1000L, 600L, 1000L, 300L, 1000L, 600L, 600L, 1000L, 1000L, 600L, 1000L, 1000L, 1000L, 1000L, 600L, 1000L, 1000L,
    1000L, 0L, 600L, 300L, 1000L, 1000L, 1000L, 100L, 1000L, 1000L, 600L, 1000L, 1000L, 1000L, 600L, 300L, 600L, 600L,
    100L, 600L, 1000L, 300L, 10L, 3L, 1000L, 300L, 300L, 1000L, 300L, 300L, 10L, 1000L, 1000L, 10L, 10L, 600L, 3L, 10L,
    600L, 600L, 600L, 1000L, 100L, 1000L, 1000L, 10L, 1000L, 3L, 10L, 1000L, 100L, 1000L, 100L, 10L, 300L, 10L, 100L,
    1000L, 0L, 1000L, 100L, 3L, 10L, 100L, 100L, 300L, 1000L, 1000L, 1000L, 10L, 1000L, 3L, 3L, 600L, 600L, 10L, 3L,
    600L, 600L, 300L, 300L, 1000L, 3L, 3L, 1000L, 10L, 1000L, 1000L, 600L, 100L, 300L, 600L, 10L, 100L, 0L, 100L, 3L,
    0L, 10L, 3L, 600L, 300L, 300L, 300L, 600L, 300L, 100L, 3L, 0L, 10L, 600L, 300L, 10L, 300L, 600L, 1000L, 600L, 600L,
    0L, 0L, 600L, 600L, 600L, 0L, 0L, 300L, 100L, 0L, 10L, 300L, 1000L, 300L, 600L, 600L, 300L, 10L, 600L, 100L, 100L,
    300L, 3L, 3L, 300L, 1000L, 10L, 3L, 100L, 3L, 100L, 100L, 300L, 3L, 3L, 600L, 300L, 3L, 3L, 3L, 300L, 3L, 0L, 10L,
    3L, 300L, 10L, 10L, 600L, 0L, 300L, 600L, 0L, 0L, 100L, 100L, 10L, 100L, 10L, 100L, 600L, 0L, 600L, 0L, 10L, 100L,
    0L, 0L, 0L, 0L, 3L, 10L, 3L, 300L, 600L, 0L, 0L, 300L, 10L, 10L, 100L, 300L, 0L, 0L, 0L, 3L, 0L, 0L, 10L, 0L, 300L,
    3L, 0L, 0L, 0L, 0L, 100L, 3L, 0L, 3L, 10L, 10L, 3L, 0L, 3L, 10L, 3L, 0L, 0L, 3L, 100L, 0L, 0L, 300L, 0L, 0L, 0L,
    0L, 0L, 0L, 0L, 3L, 0L, 3L, 0L, 0L, 0L, 0L),
  PARAM = structure(
    c(
      1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L,
      1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L,
      1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L,
      1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L,
      1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L,
      1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L,
      1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L,
      1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L,
      1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L,
      1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L,
      1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L),
    levels = "Tolerated amount of peanut protein (mg)", class = "factor"),
  USUBJID = c(
    129L, 103L, 104L, 185L, 34L, 15L, 140L, 151L, 7L, 144L, 200L, 23L, 138L, 115L, 255L, 147L, 224L, 101L, 156L, 284L,
    136L, 248L, 179L, 298L, 168L, 295L, 289L, 241L, 36L, 27L, 123L, 266L, 11L, 291L, 236L, 130L, 173L, 195L, 203L, 160L,
    274L, 167L, 290L, 94L, 9L, 269L, 122L, 135L, 64L, 26L, 95L, 10L, 5L, 234L, 161L, 299L, 88L, 69L, 35L, 233L, 286L,
    85L, 91L, 189L, 80L, 152L, 223L, 287L, 244L, 57L, 108L, 18L, 62L, 157L, 300L, 283L, 164L, 243L, 89L, 220L, 24L, 271L,
    166L, 118L, 201L, 127L, 121L, 41L, 267L, 213L, 49L, 73L, 202L, 134L, 112L, 25L, 227L, 29L, 251L, 273L, 119L, 132L,
    74L, 270L, 83L, 37L, 181L, 258L, 253L, 48L, 120L, 54L, 277L, 176L, 65L, 264L, 107L, 171L, 262L, 162L, 187L, 272L,
    288L, 294L, 245L, 109L, 172L, 204L, 275L, 22L, 66L, 186L, 247L, 17L, 149L, 141L, 177L, 280L, 216L, 40L, 75L, 263L,
    246L, 14L, 81L, 260L, 153L, 45L, 237L, 8L, 117L, 86L, 296L, 146L, 154L, 116L, 38L, 100L, 191L, 175L, 92L, 158L, 192L,
    180L, 256L, 254L, 125L, 222L, 145L, 261L, 155L, 159L, 3L, 206L, 278L, 19L, 31L, 63L, 208L, 55L, 259L, 218L, 111L,
    226L, 33L, 2L, 44L, 297L, 53L, 87L, 16L, 28L, 90L, 207L, 56L, 137L, 128L, 178L, 142L, 143L, 148L, 193L, 229L, 265L,
    97L, 252L, 205L, 150L, 165L, 188L, 52L, 99L, 93L, 1L, 221L, 124L, 210L, 6L, 232L, 21L, 211L, 163L, 96L, 60L, 183L,
    190L, 242L, 42L, 46L, 67L, 126L, 209L, 72L, 194L, 238L, 184L, 39L, 105L, 249L, 61L, 113L, 30L, 77L, 12L, 4L, 51L,
    139L, 20L, 268L, 215L, 292L, 217L, 199L, 32L, 276L, 47L, 225L, 230L, 79L, 71L, 98L, 50L, 13L, 76L, 231L, 250L, 58L,
    68L, 239L, 198L, 293L, 212L, 110L, 59L, 182L, 133L, 170L, 43L, 282L, 281L, 131L, 114L, 196L, 214L, 285L, 70L, 102L,
    279L, 106L, 197L, 257L, 228L, 84L, 235L, 78L, 169L, 240L, 219L, 174L, 82L),
  CRIT1 = structure(c(
    1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L,
    1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L,
    1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L,
    1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L,
    1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L,
    1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L,
    1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L,
    1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L,
    1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L,
    1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L,
    1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L),
    levels = "Tolerating >=600 mg of peanut protein without dose-limiting symptoms", class = "factor"),
  CRIT1FL = structure(c(
    2L, 1L, 1L, 2L, 2L, 1L, 1L, 1L, 1L, 1L, 2L, 1L, 1L, 2L, 2L, 2L, 1L, 1L, 2L, 1L, 2L, 2L, 1L, 2L, 2L, 2L, 2L, 2L,
    2L, 1L, 2L, 2L, 2L, 2L, 2L, 2L, 1L, 2L, 2L, 2L, 1L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 1L,
    2L, 1L, 2L, 2L, 2L, 1L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 1L, 2L, 2L, 1L, 2L, 2L, 1L, 1L, 1L, 2L, 1L, 1L, 2L, 1L, 1L,
    1L, 2L, 2L, 1L, 1L, 2L, 1L, 1L, 2L, 2L, 2L, 2L, 1L, 2L, 2L, 1L, 2L, 1L, 1L, 2L, 1L, 2L, 1L, 1L, 1L, 1L, 1L, 2L,
    1L, 2L, 1L, 1L, 1L, 1L, 1L, 1L, 2L, 2L, 2L, 1L, 2L, 1L, 1L, 2L, 2L, 1L, 1L, 2L, 2L, 1L, 1L, 2L, 1L, 1L, 2L, 1L,
    2L, 2L, 2L, 1L, 1L, 2L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 2L, 1L, 1L, 1L, 2L, 1L, 1L, 1L, 1L, 1L, 2L, 1L, 1L, 1L,
    2L, 2L, 2L, 2L, 1L, 1L, 2L, 2L, 2L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 2L, 1L, 2L, 2L, 1L, 1L, 2L, 1L, 1L, 1L, 1L, 1L,
    1L, 2L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 2L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 2L, 1L, 1L, 2L,
    1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 2L, 1L, 2L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 2L, 1L, 1L, 1L, 1L, 1L,
    1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L,
    1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L),
    levels = c("N", "Y"), class = "factor"),
  CRIT2 = structure(c(
    1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L,
    1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L,
    1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L,
    1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L,
    1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L,
    1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L,
    1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L,
    1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L,
    1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L,
    1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L,
    1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L),
    levels = "Tolerating >=1000 mg of peanut protein without dose-limiting symptoms", class = "factor"),
  CRIT2FL = structure(c(
    2L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 2L, 1L, 1L, 2L, 2L, 2L, 1L, 1L, 1L, 1L, 2L, 2L, 1L, 2L, 1L, 1L, 2L, 2L,
    2L, 1L, 1L, 2L, 2L, 2L, 2L, 2L, 1L, 2L, 1L, 2L, 1L, 2L, 1L, 1L, 2L, 2L, 1L, 2L, 2L, 2L, 2L, 1L, 2L, 2L, 2L, 1L,
    1L, 1L, 2L, 2L, 2L, 1L, 2L, 2L, 1L, 2L, 2L, 2L, 1L, 1L, 1L, 1L, 1L, 1L, 2L, 1L, 1L, 1L, 2L, 1L, 1L, 2L, 1L, 1L,
    1L, 2L, 2L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 2L, 1L, 2L, 2L, 1L, 2L, 1L, 1L, 2L, 1L, 2L, 1L, 1L, 1L, 1L, 1L, 2L,
    1L, 2L, 1L, 1L, 1L, 1L, 1L, 1L, 2L, 2L, 2L, 1L, 2L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 2L, 1L, 1L, 2L, 1L,
    2L, 2L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L,
    1L, 2L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 2L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L,
    1L, 2L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L,
    1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L,
    1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L,
    1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L),
    levels = c("N", "Y"), class = "factor" ) )

peanut %>%
  ggplot(aes(x=factor(AVAL), fill=TRT01P)) +
  geom_bar( position=position_dodge(preserve = "single")) +
  scale_fill_manual(values=c("royalblue", "#fb6a4a", "#ef3b2c", "#cb181d", "#a50f15", "#67000d")) +
  xlab("Peanut protein tolerated without\ndose-limiting symptoms (mg)") +
  ylab("Patients") +
  theme(legend.position="right")

peanut %>%
  group_by(TRT01P) %>%
  summarize(proportion = sum(CRIT1FL=="Y") / n()) %>%
  ggplot(aes(x=TRT01P, y=proportion, fill=TRT01P)) +
  geom_bar(stat="identity", position=position_dodge()) +
  geom_text(aes(label=scales::percent(proportion)), size=7, vjust=c(0, rep(1.5, 5))) +
  scale_fill_manual(values=c("royalblue", "#fb6a4a", "#ef3b2c", "#cb181d", "#a50f15", "#67000d")) +
  scale_y_continuous(breaks=seq(0, 0.7, 0.1), labels=scales::percent) +
  xlab("Tolerating >=600 mg of peanut protein\nwithout dose-limiting symptoms") +
  ylab("Percentage of patients")
```

#### Excercise 1A

Fit a sigmoid Emax logistic regression model 
(i.e. `family=binomial(link="logit")`). 
To accelerate likelihood evaluations, first summarize data as "responders" `y`
out of `n` patients per dose. We tell `brms` to use this information using 
`y | trials(n) ~ ...`. 
We expect the true placebo proportion to be around 0.05 (-2.944 on the logit 
scale) with much more than 0.1 or much less than 0.02 considered unlikely.
It is a-priori at least possible that there is a huge treatment effect such 
as 95\% versus 5\% responders (difference on the log-scale close to 6),
but we are at least mildly skeptical.
The dose response is a-priori somewhat likely to follow a Emax curve (with Hill
parameter near 1), but we wish to allow for the possibility of a steeper or 
shallower curve. The dose with half the effect (ED50) might be near 15 mg, but 
have considerable uncertainty around that.

```{r peanutmodel1}
model1 <- bf( y | trials(n) ~ E0 + Emax * dose^h/(dose^h + ED50^h),
              family=binomial(link="logit"),
              nlf(h ~ exp(logh)), nlf(ED50 ~ exp(logED50)),
              E0 ~ 1, logED50 ~ 1, logh ~ 1, Emax ~ 1, 
              nl=TRUE)

priors1 <- prior(normal(-2.944, 0.38), nlpar="E0") +
           prior(normal(0, 1), nlpar="logh") +
           prior(normal(0, 6), nlpar="Emax") +
           prior(normal(2.7, 2.5), nlpar="logED50")
```

Now fit the model
```{r fitpeanutmodel1}
brmfit1 <- brm(
  formula = model1,
  prior = priors1, 
  data = peanut %>% group_by(TRT01P, dose) %>% summarize(y=sum(CRIT1FL=="Y"), n=n(), .groups="drop")
)

summary(brmfit1)
```

Now use 
`tibble(dose = seq(0, 300, 1), n=1) %>% tidybayes::add_epred_rvars(object=brmfit1, newdata=.)` 
to obtain predicted proportions for every dose from 0 to 300 mg.
Then, plot the curve of predicted proportions for each of these doses using 
`ggplot2` and `ggdist` (using `ydist=.epred` in the aesthetics and the 
`stat_lineribbon()` geom from the `ggdist` package).

```{r plotmodel1preds}
tibble(dose = seq(0, 300, 1), n=1) %>% 
  tidybayes::add_epred_rvars(object=brmfit1, newdata=.) %>%
  ggplot(aes(x=dose, ydist=.epred)) +
  stat_lineribbon() +
  scale_fill_brewer(palette="Blues") +
  ylab("Model predicted proportion") +
  xlab("Dose [mg]")
```

If you prefer, replace the default fill colors e.g. with shades of blue
using `+ scale_fill_brewer(palette="Blues")

Next obtain the posterior distribution for the difference in proportions for
each of these dose levels compared with placebo.

```{r model1diffs}
tibble(dose = seq(0, 300, 1), n=1) %>% 
  tidybayes::add_epred_rvars(object=brmfit1, newdata=.) %>%
  (\(x) x %>% 
     left_join(x %>% 
                 filter(dose==0) %>% 
                 dplyr::select(-dose) %>% 
                 rename(.pbo=.epred), 
               by="n"))() %>%
  mutate(diff = .epred - .pbo) %>%
  ggplot(aes(x=dose, ydist=diff)) +
  stat_lineribbon() +
  scale_fill_brewer(palette="Blues") +
  ylab("Model predicted difference\nin proportion vs. placebo") +
  xlab("Dose [mg]")
```

#### Excercise 1B* (optional)

Try to perform a more efficient statistical analysis by either using an ordinal
outcome or treating the data as interval censored.

Try treating the data about the logarithm of the amount of protein 
tolerated as interval censored 
($-\infty$ to <log(3), >= log(3) to <log(10) mg, ...) using
`logAVAL | cens("interval", logAVALnext) ~` with `family = gaussian()`.
For the latter approach you might have to set `log(0)` to a very low number
such as -28 (approximate log(weight) of one single peanut protein 
molecule) and assume the upper interval end when a patient could tolerate
1000 mg to, say, `log(30000)` (approximately 100 peanut kernels). This
achieves approximately the same thing as setting these to `-Inf` and `Inf`,
but is handled better by `brms`/`Stan`.

How would you change the prior distributions in this case?

```{r lognormal}
model2 <- bf( logAVAL | cens("interval", logAVALnext) ~ E0 + Emax * dose^h/(dose^h + ED50^h),
              family = gaussian(),
              nlf(h ~ exp(logh)), nlf(ED50 ~ exp(logED50)),
              E0 ~ 1, logED50 ~ 1, logh ~ 1, Emax ~ 1, 
              nl=TRUE )

priors2 <- prior(normal(0, 5), nlpar="E0") + # centered on 1 mg, but very uncertain
           prior(normal(0, 1), nlpar="logh") + # No particular reason to change prior (dose response could be just as steep here as for binomial outcome)
           prior(normal(0, 6.5), nlpar="Emax") + # decent prior probability for >= 1000 mg, still conceivable could reach 300000
           prior(normal(2.7, 2.5), nlpar="logED50") + # No particular reason to change prior here (ED50 could be same as for binomial outcome)
           prior(lognormal(5, 10), class=sigma) # Additional parameter: residual SD, clearly a decent amount of variability, want to allow large SD on the log-scale

brmfit2 <- brm(
  formula = model2,
  prior = priors2,
  data = peanut %>% 
    mutate(logAVAL = ifelse(AVAL==0, -28, log(AVAL)),
           logAVALnext = case_when(AVAL==0 ~ log(3),
                                   AVAL==3 ~ log(10),
                                   AVAL==10 ~ log(100),
                                   AVAL==100 ~ log(300),
                                   AVAL==300 ~ log(600),
                                   AVAL==600 ~ log(1000),
                                   TRUE ~ log(30000)))
  )

summary(brmfit2)
```


Now use an ordinal
outcome via `family = cumulative(link = "logit", threshold = "flexible")`.
Details of this model are described in 
[a journal article](https://doi.org/10.1177/2515245918823199), for which there 
is also [a PsyArXiv preprint](https://doi.org/10.31234/osf.io/x8swp) in case
you do not have access to the journal. 

Note that the placebo response is described by a series of $K-1$ ordered 
thresholds $\theta_1, \ldots, \theta_{K-1}$ on the logit scale for $K$ ordered 
categories. For the placebo group, the probability of being in categories 
$k=1, \ldots, K-1$ is be given by $\text{logit}^{-1} \theta_k$ and for category 
$K$ by $1-\text{logit}^{-1} \theta_{K-1}$. In this case, these parameters will
appear in `summary(brmfit3)` as `Intercept[1]`, ..., `Intercept[6]`.

How would you change the model and the prior distributions in this case?

```{r ordinal}
# Note that we omit E0 from the model, because the exepcted placebo outcome
# is already described by the intercept parameters of the distribution
model3 <- bf( AVAL ~ Emax * dose^h/(dose^h + ED50^h),
              family = cumulative(link = "logit", threshold = "flexible"),
              nlf(h ~ exp(logh)), nlf(ED50 ~ exp(logED50)),
              logED50 ~ 1, logh ~ 1, Emax ~ 1, 
              nl=TRUE)

priors3 <- prior(normal(0, 1), nlpar="logh") + # No particular reason to change prior here
           prior(normal(0, 6), nlpar="Emax") + # Prior could be different (odds from one category to the next not necessarily the same as for top two vs. rest)
           prior(normal(2.7, 2.5), nlpar="logED50") + # No particular reason to change prior here
           prior(student_t(3, -0.2, 1), class=Intercept, coef=1) +
           prior(student_t(3, 0, 2.5), class=Intercept, coef=2) +
           prior(student_t(3, 0, 2.5), class=Intercept, coef=3) +
           prior(student_t(3, 0, 2.5), class=Intercept, coef=4) +
           prior(student_t(3, 2.197, 1), class=Intercept, coef=5) +
           prior(student_t(3, 0, 2.5), class=Intercept, coef=6)

brmfit3 <- brm(
  formula = model3,
  prior = priors3,
  data = peanut %>% mutate(AVAL = ordered(AVAL))
  )

summary(brmfit3)
```

For both the ordinal model and the interval-censored model, if we had a baseline 
amount of protein tolerated, we could treat this as a monotonic covariate e.g. 
using `mo(BASE)` or assume a particular functional form.