---
author:
  - Björn Holzhauer - <bjoern.holzhauer@novartis.com>
  - Andrew Bean - <andrew.bean@novartis.com>
---

# Network meta-analysis {#sec-network-meta}

This case study covers:

* Using `brms` to perform arm-based network meta-analysis using a "two-way mixed model" approach (as opposed to the also popular contrast-based or baseline-contrast model)
* Visualizing networks of evidence using function from the `multinma` package
* Setting up over-parameterized models to avoid specifying prior distributions for differences relative to a reference study or treatment (by collating two design matrices created using `model.matrix()`), and how to estimate desired contrasts in this context
* ` r | trials(n) ~ ...` for binomial outcomes (`r` out of `n` patients)
* Ranking treatments based on samples from the posterior distribution

To run the R code of this section please ensure to load these
libraries and options first:

```{r, eval=TRUE,echo=TRUE,message=FALSE,warning=FALSE}
library(tidyverse)
library(brms)
library(posterior)
library(here)
library(knitr)
library(gt)
library(netmeta) # for graphing the network & other NMA utils
# instruct brms to use cmdstanr as backend and cache all Stan binaries
options(brms.backend="cmdstanr", cmdstanr_write_stan_file_dir=here("_brms-cache"))
# create cache directory if not yet available
dir.create(here("_brms-cache"), FALSE)
set.seed(5675798)
# allow for wider printouts without line-breaking
options(width=120)

control_args <- list(adapt_delta=0.95)
```

```{r, include=FALSE, echo=FALSE, eval=TRUE}
# invisible to the reader additional setup steps, which are optional
# {{< include setup.R >}}
source("setup.R")
```



::: {.content-visible when-profile="dummy"}
## Overview Video

{{< video videos/brms-20230426-4-network-meta.mp4 >}}
:::


### Background: Why network meta-analysis?

Network (or multi-treatment) meta-analysis (NMA) is a term for analyses
that combine the results of studies that compared several treatments and
in which not all studies included all comparisons [@LuAdes2004]. This is a
frequent situation, because new drugs are often only compared to a placebo or to
one commonly used current standard of care, but not to all existing
current treatment options. As a result, physicians that want to choose
between two treatment options often find that there are no or very few
studies directly comparing them. NMA tries to address this issue by
combining evidence from direct comparisons within each trial with
indirect chains of comparisons across multiple trials.

Network meta-analyses have their limitations. One particular concern
would be factors that modify the effects of treatments resulting in
treatment by trial interactions, e.g. if pathogens developing resistance
to some drugs over time, if the presence of some background therapies
has a synergistic or antagonistic interaction with a drug, or if some
drugs work better or worse depending on disease severity.


## Data

We will use smoking cessation data [@Hasselblad1998], which consists
of the number of patients $r$ amongst the total in an study arm $n$ that
manage to stop smoking. There are 24 studies, in which different
combinations of 4 interventions (no contact, self-help, individual
counseling and group counseling) were tested. The aim of a network
meta-analysis here would be to compare these different treatment in
terms of the odds of patients managing to stop smoking.

```{r smoking_cessation_data}
data("smokingcessation", package = "netmeta")

recode_trt <- c("A" = "No_intervention",
                "B" = "Self_help",
                "C" = "Individual_counselling",
                "D" = "Group_counselling")

smoking <- smokingcessation %>%
  mutate(studyn = 1:n()) %>%
  pivot_longer(-studyn,
               names_to = c(".value", "trtid"),
               names_pattern = "(.*)([1-9])") %>%
  filter(!is.na(n)) %>%
  transmute(
    study = factor(studyn),
    trtc = factor(recode_trt[treat], unname(recode_trt)),
    trtn = as.numeric(trtc),
    r = event,
    n
  )

gt_preview(smoking)
```

```{r}
smoking %>%
  ggplot(aes(x=study, y=r/n, fill=trtc, group=trtc,
             ymin=qbeta(p=0.025, shape1=r, shape2=1+n-r),
             ymax=qbeta(p=0.975, shape1=r+1, shape2=n-r))) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_errorbar(position = "dodge", alpha=0.3) +
  theme(legend.position="bottom")
```

The totality of evidence about the treatments and their relative efficacy is 
commonly known as the *evidence network*. A common graphical representation, 
shown below, visualizes this network of treatments (the nodes) along with
edges indicating the presence or absence of direct evidence comparing pairs 
of treatments from randomized trials. Below, the thickness of the edges 
is determined by the number of trials comparing each pair of treatments.

```{r}
nm <- netmeta(pairwise(treat = trtc, event = r, n = n, studlab = study,
                       data = smoking))
netgraph(nm)
```

## Model description

We will first describe the arm-based (aka "two-way mixed model")
approach to NMA advocated by
Piepho [@Piepho2012]. It has some
advantages over more commonly used NMA approach that analyzes
treatment contrasts [@LuAdes2004], in terms of ease of use and familiarity (the
two-way linear mixed model is of course very familiar). The latter approach,
however, is widely used in health technology assessments, for example by the UK 
NICE. The NICE decision support unit has published methodological guidance on 
how to use this method in practice [@dias2014]. 



Suppose we had individual patient data for patients $k=1,\ldots,N_{ij}$ 
randomly assigned to treatment $j$ in trial $i$, for $i = 1,\ldots, n$ and
$j = 1,\ldots, m$. A two-way linear mixed model would model the outcome of
interest $Y_{ijk}$ as
$$ g(E(Y_{ijk})) = \underbrace{\alpha_i}_{\text{fixed main effect of trial }i} + \underbrace{\beta_j}_{\text{fixed main effect of treatment }j} + \underbrace{u_{ij}}_{\text{random effect of trial on treatment}},$$
where $g$ is a link function such as the logit link in case of bernoulli data.

The random effects $u_{ij}$ have mean zero and describe that the effect
of treatments is allowed to vary across trials. If we wanted to assume
the same effect of all treatment across all trials, we could instead
omit this term. When we include it, we will ideally want to allow for
the possibility that if one treatment has better outcomes in a trial,
then other treatments might also. To achieve that we assume that for the
trials $i=1,\ldots,n$ there is a multivariate normal random effect
$\boldsymbol{U}_i \sim \text{MVN}(\boldsymbol{0}, \boldsymbol{\Sigma})$,
where $\Sigma$ is some covariance matrix that is either unstructured or
has some particular structure (e.g. compound symmetric). In other words,
the components of the random vector $\boldsymbol{U}_i$ for a trial $i$
are correlated.


### Network meta-analysis with summaries for each arm

In case we have summary information by treatment arm, we will (just as
for traditional arm-based meta-analyses) often be able to assume a
normal sampling distribution for least squares means for each arm from
linear models, log-odds, log-rates or log-hazard rates. We will denote
these estimates for each arm by $\hat{\theta}_{ij}$ and their standard
error by $\text{SE}(\hat{\theta}_{ij})$. We then assume
$\theta_{ij} = \alpha_i + \beta_j + u_{ij}$ as above and that
$\hat{\theta}_{ij} \sim N(\theta_{ij}, \text{SE}(\hat{\theta}_{ij}))$.

Alternatively, for a binary outcome, we can use that the number of
patients with a success follows a binomial distribution and we would use
a regression equation like above for the logit-probability
$\theta_{ij} = \alpha_i + \beta_j + u_{ij}$ for each arm.

TO BE DETERMINE WHAT TO DO FOR SURVIAL: Less clear, could one work with
log-cumulative-hazard function (under a parametric model?)? Maybe one
could get that from KM curves? Log-(or logit-?)survival probabilities at 
fixed points of time (e.g. 1-year)? Discussed by some papers such as 
this one [@Combescure_2012]. Perhaps models fitted to 
reconstructed patient-level data/KM curves via, standard parametric models,
fractional polynomial models and/or spline-based models? What about 
non-proportional hazards?

### Patient-, arm- or trial-level covariates

As needed, the regression equation $\alpha_i + \beta_j + u_{ij}$ can be
extended by allowing for patient level covariates $\boldsymbol{x}_{ijk}$
(in case we work with individual patient data), arm-level covariates
$\boldsymbol{x}_{ij}$ or study-level covariates $\boldsymbol{x}_i$.


### Borrowing information from real-world data or other trials

If we have a fixed effect for the $\alpha_i$, then borrowing information
between trials runs into the identifiability issues between the
$\alpha_i$ and the $u_{ij}$. In a frequentist approach we cannot resolve
these issues. When we take a Bayesian approach, this identifiability
issue will still severely limit how much information we can obtain about
the distribution of the $u_{ij}$ (while we would by definition not
borrow information about the $\alpha_i$) and the results will be
sensitive to the choice of prior distributions.

We could borrow information across trials, we replace the trial main
effect by an intercept term and a random trial effect on the intercept.
However, then we will be conducting a meta-analysis, in which the causal
effects of treatments are no longer just judged based on randomized
within-trial comparisons. We would then also to an extent do comparisons
across treatment arms from different trials. Besides the limitations any
NMA is subject to, such comparisons are additionally affected by any
prognostic factors that change the expected outcomes for a treatment in
different trials. Such prognostic factors are much more common than effect
modifiers that would cause trial by treatment interactions, and we know that
to some extent such differences across trials will exist. Examples
include improving (or worsening) patient outcomes due to better
treatments or access to healthcare, different inclusion/exclusion
criteria of different trials, and differences in recruited patient
populations in terms of e.g. background therapy, disease severity,
country/region and comorbidities. While we may hope that a random trial
effect on the intercept would capture random variation in these aspects
across trials, we have to be concerned that there could be a systematic
bias in favor of some treatments in a NMA - especially if studies for
some treatments were conducted later in time than for other treatments.


## Implementation

In this case, we have a binomial outcome. 
`brms` provides the `y | trials(n)` notation to indicate `y` successes out of 
`n` trials, while in `stats::glm` the way of specifying this is a little less 
intuitive: `glm( cbind(y, n-y)~... )`. 

We will now try implement the network meta-analysis model we described
above. In the process, we will discover some things that did not make a
difference in a frequentist setting, but matter in the Bayesian setting
with respect to how we can the most easily specify our prior knowledge
(or lack thereof).

### Reference categories matter for setting priors

Firstly, before we even think too much about any prior distributions, let see how `brms` 
parameterizes the model, if we specify it in the most obvious way by just 
writing `0 + study + trtc + (0 + trtc |study)`:

```{r, results='hide', message=FALSE, warning=FALSE}
brmfit <- brm(
  data = smoking,
  formula = r | trials(n) ~ 0 + study + trtc + (0 + trtc || study),
  family = binomial(),
  control = control_args,
  prior = prior(class = b, normal(0, 3.14)),
  silent = 2,
  refresh = 0
)
```

As we can see from the summary of the fitted model, there are coefficients for
all study main effects, but not for all treatments: treatment 1 is omitted, and
the coefficients for remaining treatments now represent contrasts against treatment
1. (Conversely, if we had instead written 
`0 + trtc + study + (0+trtc|study)`, then there would have been a 
coefficient for all treatments, but the study effects would have been 
parametrized in terms of contrasts with study 1.

```{r}
summary(brmfit)
```

Is this a problem? In the frequentist context, this issue with the form of the
design matrix for population-level effects is not important, because the pairwise treatment contrasts are uniquely estimable regardless of whether constraints such as $\beta_1 = 0$ are employed.

However, if we wanted to imply some prior information or lack thereof, it may
impact our preference for the specification of the population-level terms. For
example, if we had some prior information (external to the likelihood) on a 
particular contrast between a pair of treatments, we would favor the parametrization
above, in which the treatment terms are parametrized in terms of contrasts.

### Setting priors is easier in an over-parameterized model

Another possibility is to over-parametrize the model and have a main effect for each study 
(i.e. main effects for all studies `study1` to `study24` are present) and each 
treatment. The main rationale for this would be symmetry in how we
set our prior distributions. I.e. we do not want to set our priors in a
way that would somehow favor one treatment (or study - which will have
studied some subset of treatments = might again favor somehow treatment)
over any of the others. Unlike the specification above, the marginal variance
of the each arm of each study will be constant in this approach.

This can be achieved by creating dummy variables for each treatment group and 
thereby enforcing the paramterization. 
We then specify our model as 
`0 + study + trtcA + trtcB + trtcC + trtcD + (0 + trtc | study)`.
The code below constructs the dummy variables and corresponding formula in an 
automated way. 

```{r piepho_model2, results='hide', message=FALSE, warning=FALSE}
B <- model.matrix(~ 0 + trtc, data = smoking)
S <- model.matrix(~ 0 + study, data = smoking)

smoking_with_dummies <- bind_cols(
  dplyr::select(smoking, r, n, study, trtc),
  as_tibble(B),
  as_tibble(S)
)

f <- as.formula(paste(
  "r | trials(n) ~ 0 + study +",
  paste(colnames(B), collapse = " + "),
  "+ (0 + trtc || study)"
))

# could also do:
# f <- as.formula(paste(
#     "r | trials(n) ~ 0 + trtc +",
#     paste(colnames(S), collapse = " + "),
#     "+ (0 + trtc || study)"
# ))

brmfit_with_dummies <- brm(data = smoking_with_dummies,
                           formula = f,
                           family = binomial(),
                           control = control_args,
                           prior = prior(class=b, normal(0, 3.14)),
                           silent = 2,
                           refresh = 0)
```

When we now summarize the model fit, we have regression coefficients for all 
treatment groups and studies.

```{r piepho_results2}
summary(brmfit_with_dummies)
```

While the coefficients for each treatment have changed compared to the previous
section, we should not actually look at the coefficients for each treament main
effect, but at their contrasts. When we look at those, we see that the models
lead to very similar inference (despite us not even setting any priors for 
Model 0).

```{r}
# utility function to estimate all pairwise contrasts between the main effects
# for the variable named "trtc"
contrast_draws <- function(brmfit, variable = "trtc"){
  
  trts <- levels(brmfit$data[[variable]])
  ntrt <- nlevels(brmfit$data[[variable]])
  
  all_pairs <- combn(1:ntrt, 2)
  L_trt <- t(apply(
    all_pairs, 2, function(x){
      out <- numeric(ntrt)
      out[x[1]] <- 1
      out[x[2]] <- -1
      out
    }
  ))
  colnames(L_trt) <- paste0(variable, trts)
  
  X <- standata(brmfit)$X
  L_study <- matrix(0, nrow = nrow(L_trt), ncol = sum(!grepl(variable, colnames(X))))

  # if the treatment variable is not dummy coded for every level then
  # we need to drop the base category as this is masked by study
  if(ncol(L_trt) == sum(grepl(variable, colnames(X)))){
    L <- cbind(L_study, L_trt)
  } else{
    L <- cbind(L_study, L_trt[,-1])
  }
  
  colnames(L) <- colnames(X)
  
  # labels for the contrasts
  labs <- apply(combn(trts, 2), 2, paste, collapse = " vs ")
  rownames(L) <- labs
  
  B <- as_draws_matrix(brmfit, variable = "^b_", regex = TRUE)
  
  gamma_draws <- B %*% t(L)
  
  # convert to posterior object
  as_draws_matrix(gamma_draws)
  
}

summarize_contrasts <- function(brmfit, model_name = "model", variable = "trtc", ...){
  relocate(
    mutate(summarize_draws(contrast_draws(brmfit, variable = variable), ...), model = model_name),
    model
  )
}

bind_rows(
  summarize_contrasts(brmfit, "Full-rank model"),
  summarize_contrasts(brmfit_with_dummies, "Overparametrized model")
) %>%
    gt() %>%
    fmt_number(where(is.numeric), decimals=2)
```

Note also that it is hard to interpret the study main effects in these models.
We could try to make them a little more interpretable by making them represent
something like the average expected outcome across treatment groups 
(i.e. if we leave out the treatment main effects, we get a predicted outcome
that is between the predicted outcomes for the treatment groups). For this 
purpose we encode the dummy for each treatment group as $0.5$ vs. $-0.5$ 
(if it is not that treatment group). However, treatment main effect
coefficients should still be looked at in contrast to each other.

```{r piepho_model3, results='hide', message=FALSE, warning=FALSE, eval = TRUE}
smoking_with_centered_dummies <- bind_cols(
  dplyr::select(smoking, r, n, study, trtc),
  as_tibble(B - 0.5),
  as_tibble(S)
)

brmfit_with_centered_dummies <- brm(
    data = smoking_with_centered_dummies,
    formula = f,
    family = binomial(),
    control = control_args,
    prior = prior(class=b, normal(0, 3.14)),
    silent = 2,
    refresh = 0
)
```

With the relatively low-information priors we've chosen here, all three of these models produce similar results

```{r}
all_contrasts <- tibble(
  model_name = c("Full-rank model",
                 "Overparametrized model",
                 "Overparametrized model with centering"),
  brmfit = list(brmfit, brmfit_with_dummies, brmfit_with_centered_dummies)
) %>%
    mutate(contrasts = map2(brmfit, model_name, summarize_contrasts)) %>%
    select(-brmfit) %>%
    unnest(contrasts)

all_contrasts %>%
    mutate(model=str_remove_all(model, " model"),
           model=str_replace_all(model, "with", "\nwith"),
           variable=str_replace_all(variable, "vs", "\nvs\n"),
           variable=str_replace_all(variable, "counselling", "couns."),
           variable=str_replace_all(variable, "Individual", "Indiv."),
           ) %>%
    ggplot(aes(x=median, y=model, xmin=q5, xmax=q95)) +
    geom_vline(xintercept=0, color="darkred", linetype=2) +
    geom_point() +
    geom_errorbarh() +
    xlab("log-odds ratio") +
    facet_wrap(~variable)
```


### Ranking treatments and the probability of being the best treatment

Probability that each treatment is best is straightforward in any of these approaches. To illustrate:

```{r}
B <- as_draws_matrix(brmfit_with_dummies, variable = "b_trtc", regex = TRUE)
best_idx <- apply(B, 1, which.max)
trts <- levels(smoking$trtc)
best_trt <- factor(trts, levels=trts)[best_idx]
prob_best <- prop.table(table(best_trt))

gt(arrange(as.data.frame(prob_best), -Freq)) %>% fmt_percent(Freq)
```

### Detailed look at a pairwise contrast

In some situations, there may be a particular treatment contrast that is of special interest. In some such scenarios, where we have several 
This is to illustrate that by including indirect evidence, we can enhance our inference for a pairwise contrast of interest, relative to a simple pairwise meta-analysis.

The following analyses are compared, in terms of their estimates of the relative effect (log odds ratio) of individual counselling versus no intervention:

* Network meta-analysis model (`brmfit1` above)
* Pairwise meta-analysis model (currently `RBesT` is used here, but we could equally use `brms`)
* "Stratified" analysis of the contrast separately in each study, using independent beta-binomial models


```{r fig.height=7}
#| code-fold: true
#| code-summary: "Show the code"
# identify the A vs C studies
ac_studies <- smoking %>%
  filter(trtc %in% c("No_intervention", "Individual_counselling")) %>%
  add_count(study, name = "num_treatments") %>%
  filter(num_treatments == 2) %>%
  arrange(study, trtc)

# summarize the A vs C log odds ratio using Beta-Binomial models independently
# across studies and arms
arm_level_beta_binom <- ac_studies %>%
  rowwise() %>%
  mutate(p = map2(r, n, ~ rbeta(10000, r + 0.5, n - r + 0.5))) %>%
  dplyr::select(study, trtc, p) %>%
  pivot_wider(names_from = trtc, values_from = p)

study_effects <- arm_level_beta_binom %>%
  mutate(log_odds_ratio = map2(No_intervention, Individual_counselling,
                               ~ log(.x) + log(1 - .y) - log(1 - .x) - log(.y)),
         out = map(log_odds_ratio, ~ summarize_draws(matrix(.), mean, sd, ~ quantile(., probs = c(0.025, 0.5, 0.975))))) %>%
  transmute(study = factor(study), out) %>%
  unnest(out) %>%
  dplyr::select(-variable)

# pairwise meta-analysis of the log odds ratios based on summary statistics
rbest_fit <- RBesT::gMAP(cbind(mean, sd) ~ 1 | study,
                         data = study_effects,
                         tau.dist = "HalfNormal",
                         tau.prior = 1,
                         beta.prior = cbind(0,2))

# study-level treatment-effect estimates from RBesT
fitted_rbest <- bind_cols(dplyr::select(study_effects, study), fitted(rbest_fit))

# study-level treatment effect estimates from the NMA model
nd <- inner_join(smoking_with_dummies,
                 dplyr::select(ac_studies, study, trtc),
                 c("study", "trtc")) %>%
  mutate(n = 1)

lp <- posterior_epred(brmfit, newdata = nd)
lpA <- lp[,nd$trtc == "No_intervention"]
lpC <- lp[,nd$trtc == "Individual_counselling"]
lor <- log(lpA) + log(1 - lpC) - log(1 - lpA) - log(lpC)

fitted_nma <- bind_cols(
  dplyr::select(study_effects, study),
  dplyr::select(summarize_draws(lor, mean, sd, ~ quantile(., probs = c(0.025, 0.5, 0.975))), -variable)
)

# mean and MAP treatment effect estimates from RBesT
mean_rbest <- summary(rbest_fit, type="response")[c("theta.pred", "theta")] %>% 
  do.call(what = "rbind") %>%
  as_tibble(rownames = "study") %>% 
  mutate(type = "pairwise meta",
         study = c("theta_resp_pred" = "MAP", "theta_resp" = "Mean")[study])

# mean treatment effect estimate from NMA model
mean_nma <- summarize_contrasts(brmfit, variable = "trtc", model_name = "nma", mean, sd,
                                ~ quantile(., probs = c(0.025, 0.5, 0.975))) %>%
  filter(variable == "No_intervention vs Individual_counselling") %>%
  mutate(study = "Mean", type = "network meta")

all_ests <- bind_rows(
  mutate(study_effects, type = "stratified", estimate_type = "Study-level"),
  mutate(fitted_rbest, type = "pairwise meta", estimate_type = "Study-level"),
  mutate(fitted_nma, type = "network meta", estimate_type = "Study-level"),
  mutate(mean_rbest, estimate_type = "Mean"),
  mutate(mean_nma, estimate_type = "Mean")
)

ggplot(all_ests, aes(y = study, x = mean, xmin = `2.5%`, xmax = `97.5%`,
                     group = type, color = type)) +
    geom_pointrange(position = position_dodge(0.4)) +
    facet_grid(estimate_type ~ ., space = "free", scales = "free") +
    geom_vline(xintercept = 0, linetype = "dashed") +
    scale_x_continuous(breaks=seq(-5,1,by=1)) +
    coord_cartesian(xlim=c(-5,0.5)) +
    labs(x = "Log odds ratio for individual counselling vs no intervention",
         y = "Study", color = "Analysis\nmethod") +
    theme(legend.position = "bottom")
```


## Conclusion

Arm-based (network-)meta-analysis approaches compare favorably with contrast
based approaches in many situations and fit very easily into the Bayesian
regression modeling machinery provided by `brms`. The main challenge we face
is how one should parametrize the model in order to be able to specify prior 
distributions easily, for which we showed several options. 

For NMA, Bayesian approaches are very popular due to the ease with which 
one can obtain inference about various quantities of interest (e.g. ranking of
treatments, probability that a treatment is the best one) via transformations of
the MCMC samples from the posterior distribution.

TBD: Further topics might include including observational data (aka real-world 
Exploiting that some drugs are in the same class, that some arms are different
doses of the same drug, or when we have subgroup results from the same study.


## Exercises

### Excercise 1

We will use the data on the occurrence of malignancy anti-tumour necrosis 
factor (anti-TNF) drugs used in the paper by @Warren2014. We have information 
for each arm of 13 studies of three different anti-TNF drugs (etanercept, 
adalimumab and infliximab) on how many patients developed a malignancy `y` out
of the total number of patients in an arm `n`. It is common in drug development
to simplistically reduce what is in truth a time-to-event process of adverse
event occurrence into a binomial problem. In practice, we should be cautious
about this simplification, because differential drop-out could lead to biased
conclusions and we will by necessaity get variation in effect measures for
binomial outcomes across studies of different duration even if the time-to-event
distributions are identical across studies.

```{r}
antiTNF <- tibble(
  study = c("Ericson (1999)", "Ericson (1999)", "Ericson (1999)", 
            "Moreland (1999)", "Moreland (1999)", "Moreland (1999)", 
            "Genovese (2002)", "Genovese (2002)", "Genovese (2002)", 
            "Combe (2006)", "Combe (2006)", "Van der Heijde (2006)", 
            "Van der Heijde (2006)", "Weisman (2007)/Baumgartner (2004)", 
            "Weisman (2007)/Baumgartner (2004)", "Furst (2003)", "Furst (2003)", 
            "Weinblatt (2003)", "Weinblatt (2003)", "Weinblatt (2003)", 
            "Weinblatt (2003)", "Keystone (2004)", "Keystone (2004)", 
            "Van de Putte (2004)", "Van de Putte (2004)", "Van de Putte (2004)", 
            "Van de Putte (2004)", "Breedveld (2006)", "Breedveld (2006)", 
            "Maini (2004)", "Maini (2004)", "Maini (2004)", "St Clair (2004)", 
            "St Clair (2004)", "St Clair (2004)"), 
  treatment = c("Control", "Etanercept", "Etanercept", "Control", "Etanercept", 
                "Etanercept", "Control", "Etanercept", "Etanercept", "Control", 
                "Etanercept", "Control", "Etanercept", "Control", "Etanercept", 
                "Control", "Adalimumab", "Control", "Adalimumab", "Adalimumab", 
                "Adalimumab", "Control", "Adalimumab", "Control", "Adalimumab", 
                "Adalimumab", "Adalimumab", "Control", "Adalimumab", 
                "Control", "Infliximab", "Infliximab", "Control", "Infliximab", 
                "Infliximab"), 
  regimen = c(NA, "25 mg biw", "10 mg qw or 25 mg qw or 10 mg biw", NA, 
              "25 mg biw", "10 mg biw", NA, "25 mg biw", "10 mg biw", NA, 
              "25 mg biw", NA, "25 mg biw", NA, "25 mg biw", NA, "40 mg eow", 
              NA, "40 mg eow", "20 mg eow", "80 mg eow", NA, 
              "20 mg qw or 40 mg eow", NA, "20 mg qw or 40 mg eow", 
              "20 mg eow", "40 mg qw", NA, "40 mg eow", NA, "3 mg/kg q8w", 
              "3 mg/kg q4w or 10 mg/kg q8w or 10 mg/kg q4w", NA, "3 mg/kg q8w", 
              "6 mg/q8w"), 
  dose = c(NA, "Rec", "Low", NA, "Rec", "Low", NA, "Rec", "Low", NA, "Rec", NA, 
           "Rec", NA, "Rec", NA, "Rec", NA, "Rec", "Low", "High", NA, "Rec", NA, 
           "Rec", "Low", "High", NA, "Rec", NA, "Rec", "High", NA, "Rec", 
           "High"), 
  n = c(105L, 111L, 343L, 80L, 78L, 76L, 217L, 207L, 208L, 50L, 204L, 228L, 
        454L, 269L, 266L, 318L, 318L, 62L, 67L, 69L, 73L, 200L, 419L, 110L, 
        225L, 106L, 103L, 257L, 542L, 88L, 86L, 254L, 291L, 372L, 377L), 
  y = c(0L, 0L, 2L, 0L, 1L, 0L, 4L, 5L, 5L, 0L, 1L, 1L, 10L, 2L, 2L, 0L, 4L, 0L, 
        0L, 0L, 1L, 1L, 8L, 1L, 2L,  1L, 1L, 4L, 6L, 1L, 1L, 8L, 0L, 0L, 4L))

```

If we look at the data, we can see that the proportion of patients with an 
event is quite low (<4\%) across all studies and arms. I.e. we are in a rare
event setting, where a Bayesian approach with sensibly chosen priors might be 
helpful.

```{r}
antiTNF %>%
  mutate(study = factor(study),
         arm = paste0(treatment, ifelse(is.na(dose),"",paste0(" ",dose)))) %>%
  ggplot(aes(x=study, y=y/n, 
             fill=arm, 
             group=arm,
             ymin=qbeta(p=0.025, shape1=y, shape2=1+n-y),
             ymax=qbeta(p=0.975, shape1=y+1, shape2=n-y))) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_errorbar(position = "dodge", alpha=0.3) +
  theme(legend.position="bottom") +
  guides(fill=guide_legend(ncol=2)) +
  coord_flip()
```

Try fitting models of increasing complexity (slightly different to those done 
in the @Warren2014 paper, which you could also choose to reproduce):

1. A simple meta-analysis of anti-TNF therapy vs. non-anti-TNF controls. 
2. A network meta-analysis with each anti-TNF drug having a separate fixed 
treatment effects parameter, but ignoring informaiton on doses and regimens.
3. A network meta-analysis that treats all unique combinations of drug and
dose/regimen as a separate treatment.
4. A network meta-analysis that assumes that each drug may have a different 
effect on malignancy with this effect being monotonic across doses. Note
the `mo()` notation provided by `brms` to define monotonic effects. Look
into borrowing information about the maximum effect across drugs.

<!---
### Excercise 2

**It would be good to have another example that does not involve binomial data.**

--->
