---
author:
  - Björn Holzhauer - <bjoern.holzhauer@novartis.com>
  - Sebastian Weber - <sebastian.weber@novartis.com>
---

# Probability of success from a single arm trial {#sec-pos}

This case study:
  
- discusses probability of success for drug development projects (and 
introduces the related concept of assurance) when the available data come
from a single arm trial without a control group
- shows how to specify mixture priors with `brms` using helper functions from
the `RBesT` package such as `RBesT::mixnorm()` and `RBesT::mixstanvar()` (used
with the `stanvars` option of `brms::brm()`
- gives examples of how to choose prior distributions including how to obtain
useful visualizations and other helpful information using the `ggdist` and 
`distributional` packages


To run the R code of this section please ensure to load these libraries first:
  
```{r basesetup,eval=TRUE,echo=TRUE,message=FALSE,warning=FALSE,cache=FALSE}
library(tidyverse)
library(brms)
library(RBesT)
library(knitr)
library(gt)
library(ggdist) # Great ggplo2 extension package for visualizing distributions/priors/posteriors
library(distributional) # Defines distribution (that we then use in via ggdist)
library(posterior)
library(here)

# instruct brms to use cmdstanr as backend and cache all Stan binaries
options(brms.backend="cmdstanr", cmdstanr_write_stan_file_dir=here("_brms-cache"))
# create cache directory if not yet available
dir.create(here("_brms-cache"), FALSE)
set.seed(678571)

# default controls args passed to Stan
control_args <- list(adapt_delta=0.99, step_size=0.1)

# nice common ggplot2 theme
theme_set(theme_bw(base_size=18))
```

```{r, include=FALSE, echo=FALSE, eval=TRUE}
# invisible to the reader additional setup steps, which are optional
# {{< include setup.R >}}
source("setup.R")
```




::: {.content-visible when-profile="dummy"}
## Overview Video

{{< video videos/brms3-hq-alt-3-pos.mp4 >}}

:::

## Background
  
### Probability of success
 
For pharmaceutical companies it is important to realistically assess
the probability of success (PoS) for drug development
projects. Together with other information and considerations
(e.g. costs of trials, projected timelines, unmet need for new
therapies for an indication, competitive landscape, ...) this allows
for informed decision making about things such as which projects to
pursue at all and how much at-risk investment to make into each of
them.

There are a number of ways to define success for PoS calculations, but 
[at Novartis the definition of success](https://doi.org/10.1002/pst.2179) is to 
get approval meeting a target product profile (TPP). It is assumed that if the 
drug is approved with a label as good or better than the TPP targets, it will
meet an important unmet need for patients, be successful on the market and the 
sales forecasts (obtained assuming this TPP) would be achieved. 
[The Novartis PoS framework](https://doi.org/10.1002/pst.2179) combines industry
benchmark information with data on a drug and team assessments of other risks. 
This is most easily done when we have directly relevant data, e.g. with a 
Phase 2 study that is very similar (in terms of population, endpoint, duration 
and treatment groups) to the planned Phase 3 trial(s). On the other hand, this 
is harder based on a single arm study that differs from Phase 3 (or worse
without Phase 2 data).

### Assurance

[Assurance](https://doi.org/10.1002/pst.175) is a 
[popular](https://doi.org/10.1002/pst.1856) related concept to PoS. It describes
the probability of a successful trial.

### Background of the hypothetical example

A hypothetical drug called NVS101 is already approved as a treatment for
some systemic diseases. In a single arm trial in patients with a
different systemic disease that leads to anemia, most patients
that received NVS101 for 12 weeks were able to achieve an acceptable red
blood cell count without needing a blood transfusion. We are now
considering to start a pivotal randomized placebo controlled Phase III
trial, where the primary endpoint would be having an acceptable red
blood cell count without needing a blood transfusion over a 1-year
period.

We wish to assess the PoS of such a pivotal study. I.e. the probability to get 
approval meeting the TPP of an odds ratio of 8.5 (given an assumed placebo
responder rate of 15% this corresponds to a targeted 60% for the drug).

## Data

The [Novartis PoS framework](https://doi.org/10.1002/pst.2179) provides
us with a prior for the log-odds ratio of response on drug compared with
a placebo given a number of project features. For a drug that is past
its first approval (a so-called "life-cycle management" situation) such
as NVS101, a TPP target of a odds ratio of 8.5 (given an assumed placebo
responder rate of 15% this corresponds to a targeted 60% for the drug),
we obtain a mixture prior for the log-odds ratio:

```{r}
mixture_prior <- RBesT::mixnorm("Null component"=c(0.08289228, 0, 0.91992526),
                                "TPP component"=c(0.91710772, 2.14006616, 0.91992526))

mixture_prior[,] |> 
  as_tibble() |>
  mutate(Property = c("Mixture weight", 
                      "Mean of mixture component", 
                      "SD of mixture component")) |>
  relocate(Property) |>
  gt() |>
  fmt_number(decimals=3) |>
  tab_spanner(columns=c("Null component", "TPP component"), 
              label="Components of the mixture prior")
```

Novartis associates can obtain this using the Novartis internal `pos` R
package using the `pos::benchmark_prior()` function and applying the
`pos::as_rbest_mixnorm()` function to the result, which converts the
prior to a mixture prior object in the `RBesT` package.

When we plot this mixture distribution, it looks as follows:

```{r plot_mixture}
plot(mixture_prior) + 
  xlab("Log-odds ratio for response\non drug compared with placebo") +
  theme(legend.position="bottom")
```

The hypothetical single arm trial data for NVS101 are given below:

```{r}
nvs101_singlearm <- tibble(trial="NVS101 Proof of Concept",
                           treatment="NVS101",
                           `trial duration` = "12 weeks",
                           responders=8,
                           patients=10) |>
  mutate(proportion=responders/patients,
         long_study=0L)

nvs101_singlearm |>
  gt() |>
  fmt_number(columns="proportion", decimals=2)
```

Let us assume that there has been a placebo controlled Phase 3 trial in
this indication for another drug EG-999, but no other trials with the
same endpoint. The data for the hypothetical other drug are given below:

```{r}
eg999_ph3 <- tibble(trial="EG-999 Phase 3",
                    `trial duration` = "52 weeks",
                    treatment=c("EG-999", "placebo"),
                    responders=c(19,2),
                    patients=c(26, 13)) |>
  mutate(proportion=responders/patients,
         long_study=1L)

eg999_ph3 |>
  gt() |>
  fmt_number(columns="proportion", decimals=2)
```

Superficially, this extra information makes the NVS101 single arm data
look promising, because the responder proportion on NVS101 is much
higher than on placebo and even higher than on the EG-999 active
treatment group. However, this is a non-randomized across trial
comparison and it is not immediately clear how to quantify the strength
of evidence for the efficacy of NVS101.

Since we are not really interested in EG-999, the data we will use from
the EG-999 Phase 3 trial will only be the placebo group. Thus, our the
data we will use are the following:

```{r}
our_data <- bind_rows(nvs101_singlearm,
                      eg999_ph3 |> filter(treatment != "EG-999")) |>
  mutate(treatment=factor(treatment, levels=c("placebo", "NVS101")))

our_data |>
  gt() |>
  fmt_number(columns="proportion", decimals=2)
```


## Model description

### Idea

With this available information and data, what do we do to come to a
reasonable belief about the efficacy of NVS101 and the PoS of the drug
development project? One approach that is [foreseen in the Novartis PoS
framework](https://doi.org/10.1002/pst.2179) is to conduct an expert
elicitation that relies on experts making judgements about the totality
of the available evidence.

Another approach is to analyze the available data. However, we do not
have direct data on

1.  how NVS101 compares to a placebo group within the same trial (the
    proof of concept study was a single arm trial),

2.  how much the true placebo group responder proportion varies from
    trial to trial, and

3.  how much the true responder proportion changes with trial duration
    (for either NVS101 or placebo).

Thus, we will still require prior judgements on these questions.
However, it could be the case that these questions are easier to make
judgements about for experts than if we asked them directly about the
efficacy of a drug compared with placebo.

### Model specification

we first express a mixed effects logistic regression model ignoring
whether we have data to tell us something about all model parameters. We
assume a binomial outcome
$Y_{ij} \sim \text{Bin}(\pi_{ij}, N_{ij})$ for arm $j$ of trial $i$. Our
model specifies that

\begin{align}
\text{logit}(\pi_{ij}) =& \; \beta_0  \\
& + \beta_1 \times 1\{\text{treatment}_{ij} = \text{NVS101}\} \\
& + \beta_2 \times
1\{\text{duration} = \text{52 weeks} \} \\
&+ \beta_3 \times
1\{\text{treatment}_{ij} = \text{NVS101 \& duration} = \text{52 weeks}
\}  \\
& + \theta_i,
\end{align}

where $\theta_i \sim N(0, \tau^2)$ is a random trial effect. Or to
express this as a brms-formula:

```{r}
model_formula <- bf(responders | trials(patients) ~ 1 + treatment + long_study + long_study:treatment + (1 | trial),
                    # short notation for the rhs of this formula is
                    # 1 + treatment*long_study + (1 | trial)
                    center = FALSE, 
                    family = binomial(link="logit"))
```

**Note** that we use the option `center=FALSE` here in order to fully
control the parametrization of the linear predictor. By default `brms`
centers the design matrix and treats the intercept separately from the
remaining covariates. While this is often desirable to do, this is not
appropiate in this case where we want to control all details of the
model and intend to set priors deliberately.

### Specification of prior distributions

Before we saw the EG-999 Phase 3 study, what might we have believed
about the placebo group of such a study? To answer this might be a time
consuming process that may require a review of the available literature
and/or eliciting the judgements of experts. Conveniently, the sample
size section of the clinical trial protocol of the EG-999 Phase 3 study
contained a statement that the expected placebo responder rate might be
in the range of 5% to 40%. We will interpret this to indicate a 80% prediction
interval. The authors of the protocol do not really say so clearly, but
we speculate that we cannot rely on this being a 95% prediction
interval. Thus, if we assume a normal distribution for the predictive
distribution, we get a normal distribution on the logit-scale with mean
`( logit(0.4) + logit(0.05) )/2` $\approx$
`r round( ( logit(0.4) + logit(0.05) )/2, 2)` and standard deviation
`( logit(0.4) - logit(0.05) ) / 2 / qnorm(0.9)` $\approx$
`r round( ( logit(0.4) - logit(0.05) ) / 2 / qnorm(0.9), 1)`. 

::: {.alert .alert-info}
Note that the `logit` and `inv_logit` are convenient functions from the `RBesT`
package for converting from probabilities to logit-probabilities and
vice versa (in base R you can use `qlogis` and `plogis` instead).
:::

The resulting prior distribution looks as shown in the figure below.
```{r}
tibble(mean=-1.67, sd=1.0) |>
  ggplot(aes(xdist = dist_normal(mu = mean, sigma = sd))) +
  stat_halfeye(fill="darkorange", alpha=0.5) +
  scale_x_continuous(breaks=seq(-4,2,1), 
                     labels=paste0(seq(-4,2,1), "\n(=", 
                                   round(inv_logit(seq(-4,2,1)),2)*100,"%)")) +
  xlab("Average placebo log-odds") +
  ylab("Prior density")
```

::: {.alert .alert-info}
**NOTE**: The `ggdist` and `distributional` packages are an amazing
combo for understanding distributions. `ggdist` extends `ggplot2` to
help us with visualizing distributions (see
[here](https://mjskay.github.io/ggdist/) for more information).
`distributional` lets us work with distributions including many commonly
used ones, truncated distributions (e.g. half-normal as
`dist_truncated(dist_normal(mu = 0, sigma=2.5), lower=0)`),
distributions of transformed random variables (e.g.
`dist_transformed(dist=dist_normal(0, 0.5), transform=exp, inverse=log) |> mean()`)
and mixture distributions (e.g.
`dist_mixture(dist_normal(0, 1), dist_normal(1.5, 5), weights = c(0.5, 0.5))`).
With the `distributional` package, we can also obtain information on
distributions such as their mean, median, quantiles, variance, skewness
and so on (e.g. `dist_student_t(df=3, mu = 0, sigma = 1) |> median()`).
:::

Now that we have a prior for the average true response proportion, we 
consider how much the true proportion might vary from study to study. 
To set a prior distribution on the between trial SD, we asked experts
about the narrowest and widest random effects distributions
they consider plausible. 
When discussing with our experts, we fixed the 
random effects mean to a plausible value to make it 
easier for them to express their prior belief on a scale they are 
comfortable on. We then assumed that the same distribution width is 
approximately plausible for other mean values, too.

For a hypothetical scenario with an average true reponse proportion of 15\%,
clinical experts expressed that it would be unlikely that a few
studies would not end up having a true responder proportion of 20\% and
that it is possible that variability could be so high that some could
reach 40\%.

Thus, we considered that for the narrowest plausible random effects 
distribution (conditional on the the middle of the distribution being 15\%),
there should be 2.5\% probability that studies could vary to 20\% or higher.
Similarly, for the widest plausible distribution, we assigned 2.5% probability
that studies would vary to 40% or more. Normal distributions on the logit scale
that fulfill these criteria are shown in the figure below.

```{r}
tibble(mu=qlogis(0.15), sigma=c(0.18, 0.68) ) %>% 
  mutate(res=map2(mu, sigma, \(mu,sigma) tibble(x=seq(-2.5,0.5,0.001), y=dnorm(seq(-2.5,0.5,0.001), mu, sigma)))) %>%
  unnest(res) %>%
  mutate(grouping = factor(case_when(
    (sigma==0.18 & x>qlogis(0.2)) ~ 1L,
    sigma==0.18 ~ 2L,
    (x>qlogis(0.4)) ~ 3L,
    TRUE ~ 4L), levels=rev(1L:4L))) %>%
  ggplot(aes(x=x, y=y, ymin=0, ymax=y, col=grouping,  fill=grouping)) +
  theme(axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        legend.position="none") +
  geom_vline(xintercept=qlogis(0.15), lty=2, col="darkred") +
  #stat_slab(slab_alpha=0.5, n=10000) +
  geom_ribbon(alpha=0.4) +
  coord_cartesian(xlim=c(-2.2, 0)) +
  scale_x_continuous(breaks=qlogis(seq(0.05, 0.5, 0.05)),
                     labels=paste0(round(qlogis(seq(0.05, 0.5, 0.05)), 2), "\n=", 100*seq(0.05, 0.5, 0.05), "%")) +
  geom_text(aes(x=x, y=y, label=label),
             data=tibble(x=c(-1.62, -1.4, -1.1, -0.9), y=c(2, 0.8, 0.6, 0.4), 
                         label=c("SD of 0.18",
                                 paste0("P(probability>=20%)=",round(100-100*pnorm(q=qlogis(0.2), qlogis(0.15), 0.18),1), "%"),
                                 "SD of 0.68",
                                 paste0("P(probability>=40%)=",round(100-100*pnorm(q=qlogis(0.4), qlogis(0.15), 0.68),1), "%")),
                         mu=0, sigma=c(0.18, 0.18, 0.68, 0.68), grouping=factor(c(2L, 1L, 4L, 3L), levels=rev(1L:4L))), size=6, hjust = 0) +
  xlab("True logit-response-probability for new trial\ngiven average trial is -1.73 (=15%)") +
  ylab("Prior density") +
  scale_fill_manual(values=rev(c("darkred", "#E69F00", "darkblue", "#56B4E9"))) +
  scale_color_manual(values=rev(c("darkred", "#E69F00", "darkblue", "#56B4E9"))) +
  ggtitle("Narrowest & widest plausible random effects\ndistribution conditional on a mean of logit(0.15)")
```

Now, we translate these beliefs into a prior for the between-trial SD.
Given that we consider 0.18 and 0.68 as on the edge of what is plausible 
for the between-trial SD, we pick a log-normal distribution for the 
between-trial SD that has these two values approximately as its 
2.5th and 97.5th percentiles. For such a normal distribution on the logit-scale 
centered on logit(0.15), this implies a standard deviation of
`(logit(0.2)-logit(0.15)) / qnorm(0.975)` $\approx$ `r round((logit(0.2)-logit(0.15)) / qnorm(0.975),2)` to
`(logit(0.4)-logit(0.15)) / qnorm(0.975)` $\approx$ `r round((logit(0.4)-logit(0.15)) / qnorm(0.975),2)`. This can be
described by a log-normal prior distribution on the standard deviation
with location -1.06 and scale 0.35, as we can see by observing that
`dist_lognormal(mu = -1.06, sigma = 0.35) |> quantile(p=c(0.025,
0.975)) |> unlist()`
=
`r round(dist_lognormal(mu = -1.06, sigma = 0.35) |> quantile(p=c(0.025, 0.975)) |> unlist(), 3)`.

```{r}
tibble(mu=-1.06, sigma=0.35) |>
  ggplot(aes(xdist = dist_lognormal(mu=mu, sigma=sigma))) +
  geom_vline(xintercept=0.18, col="#E69F00", linewidth=2, lty=2) +
  geom_vline(xintercept=0.68, col="#56B4E9", linewidth=2, lty=2) +
  geom_text(data=tibble(mu=c(0.3, 0.8), 
                        sigma=c(0.96, 0.5), 
                        label=c("SD of 0.18", "SD of 0.68")),
            aes(x=mu, y=sigma, label=label),
            size=6, col=c("#E69F00", "#56B4E9")) +
  stat_halfeye(n=10000, fill="red", alpha=0.6) +
  coord_cartesian(xlim=c(0, 1.0)) +
  ylab("Prior density") +
  xlab("Between trial SD")
```

When asked how much lower the responder proportion over 52 vs. 12 weeks would 
be, the clinicians judge it likely that the proportion would decline a little
bit. Their rationale was that while the disease does not progress fast enough
to change substantially over 9 months, a longer duration offers additional 
opportunities for a low red blood cell count to occur (even if by chance) that
might lead to a blood transfusion.

The resulting prior distribution looks as shown in the figure below.
```{r}
tibble(mean=-0.1, sd=.1) |>
  ggplot(aes(xdist = dist_normal(mu = mean, sigma = sd))) +
  stat_halfeye(fill="#e7298a", alpha=0.5) +
  scale_x_continuous(breaks=log(1+seq(-0.3, 0.2, 0.1)), 
                    labels=paste0(abs(round(seq(-0.3, 0.2, 0.1) * 100,0)), 
                                  "%\n", 
                                  c(rep("lower\nodds", 4), rep("higher\nodds", 2)))) +
  xlab("Log-odds of response over 52 compared with 12 weeks") +
  ylab("Prior density")
```

Finally, the clinicians were unsure whether a 52 week vs. a 12 week study
would result in a higher or lower log-odds-ratio for drug vs. placebo.
However, they expressed that the effect would be relatively small, because
the drug should not be disease modifying (i.e. treatment effect should not
grow over time) and there is no reason to speculate that the effect on 
red blood cell counts would somehow decline over time, either. On this basis,
they expressed the following prior distribution:
```{r}
tibble(mean=0, sd=.1/1.96) |>
  ggplot(aes(xdist = dist_normal(mu = mean, sigma = sd))) +
  stat_halfeye(fill="royalblue", alpha=0.5) +
  scale_x_continuous(breaks=log(1+seq(-0.3, 0.2, 0.1)), 
                    labels=paste0(abs(round(seq(-0.3, 0.2, 0.1) * 100,0)), 
                                  "%\n", 
                                  c(rep("lower\nodds", 4), rep("higher\nodds", 2)))) +
  xlab("Change in log-odds ratio of drug vs. placebo\nwhen increasing trial duration from 12 to 52 weeks") +
  ylab("Prior density")
```

Thus, we can set out prior distributions as below. To see what particular
coefficients are named, `get_prior(model_formula, our_data)` is a useful
command.

```{r}
brms_prior <- prior(class=b, coef=Intercept, normal(-1.67, 1.0)) + # prior average placebo effect 
  prior(class=sd, coef=Intercept, group=trial, lognormal( -1.06, 0.35)) + # prior on the between-study SD in  placebo log-odds
  prior(class=b, coef=long_study, normal(-0.1, 0.1)) + # prior for effect of long-study (26 vs. 12 weeks) on responder rate
  prior(class=b, coef="treatmentNVS101:long_study", normal(0, 0.1/1.96)) + # Prior on the log-odds-ratio for drug vs. placebo when switching from shor to long study
  prior(mixnorm(map_w, map_m, map_s), class=b, coef=treatmentNVS101) #  Syntax for specifying mixture prior
```

::: {.alert .alert-info}
**NOTE**: Thanks to the new functionality in the `RBesT` package, we were able
to just write `prior(mixture_prior, class=b, coef=treatmentNVS10)` to specify
a mixture prior. Alternatively, we could have written
`prior(mixnorm(map_w, map_m, map_s), class=b, coef=treatmentNovDrug)`. 
`map_w`, `map_m` and `map_s` would specify the weights, means and standard
deviations of the mixture components, respectively. If we did not have this
`RBesT` functionality available, we would have to manually write down the
log-probability-density-function of the mixture, which involves weighted sums
on the probability scale e.g. using the `log_sum_exp` function. For
further details on the `mixstanvars` feature, please refer to its
[online manual](https://opensource.nibr.com/RBesT/reference/mixstanvar.html).
:::

### Model fitting and predictions

```{r fitmodel, results='hide',message=FALSE,warning=FALSE}
brmfit1 <- brm(formula = model_formula,
               data = our_data,
               # By doing 11,000 samples per chain (default 4 chains) incl. 1000
               # warmup samples, we'll get 4 * 10,000= 40,000 samples, which is
               # what we'll need in the PoS app.
               #iter = 11000, warmup=1000, 
               # Now we specify our prior distributions:
               prior = brms_prior,
               # Note that we define "map=mixture_prior" below. This
               # matches the prefix "map" defined in the previous
               # prior definition statement:
               stanvars = mixstanvar(map=mixture_prior),
               control = control_args,
               seed = 5674574,
               silent = 2, 
               refresh = 0
               )
```
We can now generate predictions for a new Phase 3 trial 
for NVS101 compared with placebo of 52 week duration. While the predictions
for each arm of the new trial are by default for the true log-odds of response,
we can derive all desired quantities from this:

```{r ph3preds}
# we are using the inv_logit function on rvars for which we need to
# prepare the funnction to handle rvar arguments
rvar_inv_logit <- rfun(inv_logit)

new_ph3_trial <- tibble(trial="New trial",
                        treatment=c("placebo", "NVS101"),
                        long_study=1,
                        patients=c(25, 50))

ph3_predictions <- new_ph3_trial |>
    tidybayes::add_linpred_rvars(brmfit1,
                                 allow_new_levels=TRUE,
                                 sample_new_levels="gaussian",
                                 value="logodds") |>
    mutate(proportion=rvar_inv_logit(logodds))

# One can have tibbles contain columns of rvars, which one can
# recognize by the reported mean and standard deviation of the
# sample they represent:
ph3_predictions

```


## Results

Here is a plot of the predicted Phase 3 placebo and NVS101 responder proportion.
```{r}
with(ph3_predictions,
     tibble(proportion_NVS101  = proportion[2],
            proportion_placebo = proportion[1])) |>
    tidybayes::unnest_rvars() |>
    ggplot(aes(x=proportion_placebo, y=proportion_NVS101)) +
    geom_hex() +
    scale_fill_continuous() +
    coord_cartesian(xlim=c(0,1), ylim=c(0,1))
```

Additionally, we can look at various measures of a treatment effect:
```{r}

ph3_predictions |>
    ggplot(aes(xdist=proportion)) +
    stat_slabinterval() +
    facet_wrap(~treatment, labeller=label_both, scales="free") +
    xlab("Proportion") + ylab(NULL)

with(ph3_predictions,
     tibble(measure=c("logodds", "lograte", "proportion_diff"),
            posterior=c(diff(logodds),
                        diff(log(proportion)),
                        diff(proportion)),
            ref=c(1,1,0)
            )) |>
    ggplot(aes(xdist=posterior)) +
    stat_slabinterval() +
    geom_vline(aes(xintercept=ref), lty=2, col="darkred") +
    facet_wrap(~measure, scales="free_x") +
    ylab(NULL)
     


```

Let's say the target product profile (TPP) for the drug demands a risk difference
of >= 45\%. We can simulate Phase III outcomes to find the probability of 
meeting this TPP and having a significant study, which the Novartis internal
PoS app would do for us.

Note that the relativley compact code below does indeed perform a full
trial simulation for possible outcomes for every draw of the
posterior. The use of `rvars` hides essentially that we are handling
entire posteriors here. However, note that we have to wrap the call to
the `fisher.test` funcntion into `rdo` such that the expression is
automatically evaluated for each draw of the posterior.

```{r}
pos_simulations <- new_ph3_trial |>
    tidybayes::add_predicted_rvars(brmfit1,
                                   allow_new_levels=TRUE,
                                   sample_new_levels="gaussian",
                                   value="pred") |>
    mutate(predrate=pred/patients) |>
    select(treatment, pred, predrate, patients) |>
    pivot_wider(names_from="treatment", values_from=c("pred", "predrate", "patients")) |>
    mutate(riskdiff_crit = predrate_NVS101 - predrate_placebo >= 0.45,
           pvalue = rdo(fisher.test(x=matrix(c(pred_NVS101,  patients_NVS101  - pred_NVS101,
                                               pred_placebo, patients_placebo - pred_placebo ),
                                             2, 2 ) )[["p.value"]] ),
           `Achieving >= 45% risk difference` = E(riskdiff_crit),
           `Achieving significance (p<=0.05)` = E(pvalue <= 0.05),
           `Achieving significance & TPP target` = E(riskdiff_crit * (pvalue <= 0.05))
           )

pos_simulations |>
    select(starts_with("Achieving")) |>
    pivot_longer(cols=everything(), names_to="Outcome", values_to = "Probability") |>
    gt() |>
    fmt_number(decimals=3)



```

As we can see, achieving the TPP risk difference is the greater hurdle than
achieving statistical significance.

As part of the Novartis PoS framework, we would now account for
the possibility of not obtaining an approval despite a positive Phase 3 study,
a failure in Phase 3 due to a safety issue, and other risks (e.g. relating
to drug manufacturing or market access).


## Conclusion

In the hypothetical case study, there was not a lot of data, but what there was 
appeared to be promising. Thus, one would not want to ignore it. 
In order to make use of it, we had to use informative prior distributions for 
several parameters of a logistic regression model. In this kind of situation,
where we cannot let data inform our prior distributions, there is little choice,
if we want an answer using the available data.

Given how our prior judgments directly affect the inference, we want ot ensure 
that the prior judgements reflect our best understanding of the
available scientific evidence. For this purpose a formal prior elicitation
(or predictions for some parameters from historical data) are options, but
when those are not possibler, we should still try to ensure that prior choices
are plausible to the whole team.

Note that in this example, there was also an underlying laboratory measurement
(longitudinal measurements of red blood cell counts). It might have been more
efficient to work with these directly, but it would have been necessary 
to translate any results into predictions for the Phase III responder endpoint.
Furthermore, the published external competitor data might not report these
laboratory outcomes in sufficient detail.

Single arm studies are not ideal. However, when treatment effects are large
compared to how much placebo group outcomes vary, we can still learn
about drug efficacy from them at the cost of additional assumptions.
As part of our case study, we tried to be explicit about our 
uncertainties and to transparently describe our prior judgements, as well as 
the rationale behind them. If someone disagrees with some of our prior 
distributions, they can then do a sensitivity analysis.

A Bayesian approach is also well-suited for PoS calculations, because it lets
us easily propagate uncertainties through different steps of a PoS analysis.

## Exercises

To be added

## References {.unnumbered}

::: {#refs}
:::
