---
author:
  - Björn Holzhauer - <bjoern.holzhauer@novartis.com>
format:
  html:
    toc: true
    number-sections: true
    html-math-method: katex
bibliography: references.bib
---

# Difference in proportions from a fitted logistic regression {#sec-diffp}

<!-- https://raw.githubusercontent.com/Novartis/bamdd/main/src/_macros.qmd -->
{{< include _macros.qmd >}}

This case study features:
  
- how to obtain a difference between proportions after
  fitting a Bayesian logistic regression using the standardized
  estimator approach
- how to use sum-to-zero contrasts to express a-priori equal
  uncertainty about multiple factor levels
- that a normal prior distribution with a standard deviation
  of 3.14 for the intercept of a logistic regression is already very
  vague
- the use of `rvars` and the convenient helper functions
  such as `tidybayes::add_epred_rvars()`
- collapsability and transportability of effect measures, as
  well as population average and individual treatment effects

```{r includestuff, include=FALSE, cache=FALSE}
here::i_am("src/02m_proportion_difference.qmd")
source(here::here("src", "setup.R"))
```

To run the R code of this section please ensure to load these libraries first:
  
```{r basesetup,eval=TRUE,echo=TRUE,message=FALSE,warning=FALSE,cache=FALSE}
library(tidyverse)
library(brms)
library(posterior)
library(tidybayes)
library(gt)
library(here)

# instruct brms to use cmdstanr as backend and cache all Stan binaries
options(brms.backend="cmdstanr", cmdstanr_write_stan_file_dir=here("_brms-cache"))
# create cache directory if not yet available
dir.create(here("_brms-cache"), FALSE)
set.seed(678571)

# default controls args passed to Stan
control_args <- list(adapt_delta=0.99, step_size=0.1)

# nice common ggplot2 theme
theme_set(theme_bw(base_size=18))
```

```{r brmscachestuff, include=FALSE, cache=FALSE}
# speedup things by lowering adapt_delta for now, which speeds up things a lot
# do this only if caching is on (which is when we want things go fast)
# TODO: Can we set adapt_delta to low values based on knowing this is a non-production run??
control_args <- list(adapt_delta=0.8)

# move caching directory to a cpu specific directory (for those using
# Docker)
options(cmdstanr_write_stan_file_dir=here("_brms-cache", Sys.info()["machine"]))
# create cache directory if not yet available
dir.create(here("_brms-cache", Sys.info()["machine"]), FALSE)

```

## Background

### Different effect measures

Statisticians will often report results of an analysis in terms of the 
parameters of their analysis model, such as a (log-)odds ratio based on a 
logistic regression for a binary outcome. However, physicians or patients may
struggle to understand these effect measures. They might instead prefer 
alternative effect measures such as a difference in proportion, the number of
patients they need to treat with a drug (instead of with placebo) before we
expect "it to make a difference for one of the patients" 
(aka "number needed to treat" or NNT) or the ratio of proportions (risk-ratio).

Even once we have chosen an effect measure, another question is whether we want
it in its marginal (population-average) or conditional (patient-specific) form.
Additionally, different effect measures may have other desirable properties such
as collapsability and transportability.

### Marginal (population-average) or conditional (patient-specific)

A patient (and their physician) will usually be interested in how much a 
treatment will help them ("How much would taking the drug reduced my risk of 
being hospitalized in the next year?") taking their specific situation into 
account. This is the **conditional** (on patient covariates) or 
**patient-specific** question.

In contrast, an insurance company or 
state healthcare system might be more interested in the effect an intervention 
has on a population level ("How much would giving the drug to all eligible 
patients reduce the number of hospitalizations in the next year?"). This is a
**marginal** or **population-average** effect question.

<div class="alert alert-info">
  The difference between conditional and marginal estimates is not whether they
  come from a covariate adjusted model. Marginal estimates can still be 
  obtained (with gain in precision) from covariate adjusted analyses.
</div>

Even when a treatment actually has the same causal effect for all
patients, the population-average effect may not be the same as the
patient specific effect.  In general, we call an effect measure
**collapsible** when the population-average effect is the average of
the effects for the members of the population. In contrast, an effect
measure, for which this is not the case is **non-collapsible**.

The difference in proportions is a **collapsible effect measure**. I.e. the 
marginal (population average) difference in proportions is the mean of the 
difference in proportions across the individuals in the population. 
This also means that, if the modelled difference in proportions is the same for 
all patients, then the marginal difference in proportions is the same as for 
each patient.
  
However, mathematically, a difference in proportions cannot be constant over 
differing control group proportions. It may just not vary to a degree we 
consider practically relevant. The difference in proportion for one patient 
will not generally be the same as for another patient and there may be no 
patient for whom the marginal difference in proportion applies (it is simply 
an average across patients). Thus, the marginal difference in 
proportions will differ for two populations with different covariate 
distributions. For that reason 
[Harrell has discussed](https://www.fharrell.com/post/rct-mimic/) that the 
marginal difference in proportions is not a **transportable** effect measure, 
while the odds ratio might be transportable (but is non-collapsible).
  
I.e. a marginal difference in proportions for a drug compared with placebo
from a clinical trial may not be a good summary of what will happen in 
clinical practice, because the distribution of characteristics of trial
participants will rarely exactly match the distribution of the real-world
patient population for the drug.
  
[The FDA guidance](https://www.fda.gov/regulatory-information/search-fda-guidance-documents/adjusting-covariates-randomized-clinical-trials-drugs-and-biological-products) 
illustrates the difference between the odds ratio and the difference in 
proportions. This example illustrates the non-collapsibility of the odds ratio 
in a hypothetical target population (the marginal odds ratio is lower than the 
odds ratio for both subgroups) and the collapsibility of the difference in 
proportions (the marginal difference in proportions is exactly the weighted 
average of the subgroup differences weighted by the percentage of the target 
population in the subgroup).
  
+-------------------+-------------------+--------------+--------------+-------+---------------+
|                   | Percentage of     | Success rate | Success rate | Odds  | Difference in |
| Population        | target population | New drug     | Placebo      | ratio | proportions   |
+-------------------+-------------------+--------------+--------------+-------+---------------+
| Biomarkerpositive | 50%               | 80.0\%       | 33.3\%       | 8.0   | 46.7\%        |
+-------------------+-------------------+--------------+--------------+-------+---------------+
| Biomarkernegative | 50%               | 25.0\%       | 4.0\%        | 8.0   | 21.0\%        |
+-------------------+-------------------+--------------+--------------+-------+---------------+
| Combined          | 100%              | 52.5\%       | 18.7\%       | 4.8   | 33.8\%        |
+-------------------+-------------------+--------------+--------------+-------+---------------+

In this example, an conditional (on covariates) odds ratio of 8.0 applies to all 
patients and would be a reasonable patient-specific effect measure.

For a population with 50\% biomarker-positive and 50\%
biomarker-negative patients a difference of proportion of 33.8\% is
the correct marginal (population-average) difference in
proportions. However, this marginal effect applies to no patient (the
difference in proportions for a patient is either 46.7\% or 21.0\%
depending on the subgroup the patient is in). It also changes when we look
at a population with, say, 80\% biomarker-negative patients, then the
marginal difference becomes 26.1\%. If the data above have arisen from
a clinical trial, in which the equal size of the biomarker-positive
and -negative subgroups was solely an effect of ensuring equal
enrolment in both subgroups, then it seems hard to justify why the
equal weighting of the subgroup-specific differences in proportions is
relevant to clinical practice. It is, of course, possible to adjust a
difference in proportions to a known covariate distribution for a
population of interest.

### Marginal difference in proportions based on a covariate-adjusted Bayesian analysis

In this case study, we illustrate how to obtain a marginal difference in 
proportions based on a covariate-adjusted Bayesian analysis, as well as from 
one not adjusted for covariates. The same approach can also be used for 
e.g. the ratio of proportions, the difference in exponential hazard rates or 
the difference in the proportion of patients with an event after 1-year based 
on a survival model, or the difference in the proportion of patients achieving 
blood pressure control based on a linear regression.

## Example data

The example concerns a hypothetical Phase IIA randomized controlled trial (RCT) 
conducted in four clinical trial sites. The RCT compared a new drug with placebo 
in terms of the proportion of patients with a disease achieving levels of a 
blood biomarker above the lower limit of normal for a healthy population. 
Younger patients are more likely to have a faster progressing and more 
aggressive form of the disease, so they are expected to be less likely to 
achieve this.

```{r thedata}
#| code-fold: true
#| code-summary: "Example data"
logit <- qlogis
inv_logit = plogis

trial_data <- tibble(
    USUBJID = 1:250, 
    TRT01P = c(
        1L, 1L, 1L, 1L, 0L, 1L, 0L, 0L, 0L, 1L, 1L, 1L, 1L, 0L, 1L, 1L,
        1L, 0L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 0L, 1L, 1L, 1L, 0L, 1L,
        0L, 0L, 0L, 1L, 0L, 1L, 1L, 0L, 1L, 0L, 1L, 1L, 1L, 0L, 0L, 0L,
        0L, 1L, 1L, 1L, 1L, 1L, 0L, 1L, 0L, 0L, 1L, 0L, 1L, 1L, 0L, 1L,
        0L, 0L, 1L, 0L, 0L, 1L, 1L, 0L, 0L, 0L, 1L, 1L, 1L, 0L, 1L, 0L,
        0L, 1L, 0L, 1L, 0L, 1L, 0L, 1L, 1L, 0L, 0L, 0L, 0L, 1L, 0L, 1L,
        1L, 0L, 1L, 1L, 1L, 0L, 0L, 0L, 0L, 1L, 0L, 1L, 0L, 0L, 0L, 1L,
        1L, 1L, 1L, 0L, 1L, 0L, 1L, 0L, 0L, 1L, 1L, 0L, 1L, 0L, 1L, 1L,
        1L, 1L, 0L, 0L, 1L, 0L, 0L, 1L, 1L, 1L, 1L, 1L, 0L, 1L, 0L, 1L,
        1L, 0L, 1L, 1L, 0L, 1L, 1L, 1L, 0L, 0L, 0L, 0L, 1L, 1L, 1L, 0L,
        0L, 0L, 0L, 1L, 1L, 1L, 1L, 1L, 0L, 1L, 1L, 0L, 0L, 0L, 0L, 1L,
        1L, 0L, 1L, 0L, 0L, 0L, 0L, 0L, 1L, 1L, 0L, 1L, 0L, 1L, 1L, 1L,
        0L, 1L, 1L, 0L, 1L, 0L, 0L, 0L, 1L, 0L, 1L, 0L, 0L, 0L, 1L, 1L,
        1L, 0L, 1L, 0L, 1L, 0L, 0L, 0L, 1L, 0L, 0L, 0L, 1L, 1L, 1L, 0L,
        0L, 0L, 0L, 1L, 0L, 1L, 1L, 0L, 0L, 0L, 0L, 1L, 0L, 1L, 1L, 1L,
        0L, 1L, 0L, 0L, 1L, 0L, 1L, 0L, 1L, 1L
    ),
    AGE = c(
        77L, 49L, 51L, 77L, 73L, 76L, 58L, 77L, 69L, 52L, 45L, 67L, 42L,
        42L, 56L, 76L, 71L, 46L, 76L, 54L, 67L, 40L, 37L, 70L, 70L, 45L,
        69L, 67L, 63L, 35L, 64L, 62L, 78L, 53L, 44L, 55L, 47L, 46L, 63L,
        65L, 67L, 79L, 44L, 56L, 47L, 61L, 74L, 52L, 57L, 65L, 36L, 58L,
        72L, 59L, 48L, 72L, 71L, 45L, 66L, 80L, 62L, 75L, 67L, 67L, 66L,
        49L, 72L, 55L, 44L, 68L, 77L, 59L, 84L, 78L, 77L, 71L, 36L, 72L,
        36L, 39L, 57L, 79L, 71L, 43L, 61L, 72L, 49L, 56L, 54L, 75L, 68L,
        45L, 58L, 46L, 39L, 65L, 73L, 72L, 71L, 49L, 65L, 70L, 57L, 47L,
        62L, 36L, 76L, 70L, 84L, 54L, 76L, 36L, 51L, 38L, 56L, 44L, 58L,
        53L, 47L, 80L, 75L, 47L, 61L, 57L, 49L, 64L, 53L, 44L, 47L, 67L,
        40L, 56L, 69L, 81L, 67L, 83L, 84L, 81L, 81L, 55L, 38L, 48L, 61L,
        80L, 51L, 76L, 82L, 61L, 44L, 51L, 48L, 72L, 56L, 49L, 73L, 71L,
        61L, 72L, 47L, 45L, 42L, 82L, 84L, 50L, 46L, 57L, 66L, 49L, 74L,
        70L, 47L, 48L, 75L, 53L, 67L, 38L, 58L, 48L, 65L, 44L, 47L, 47L,
        61L, 76L, 77L, 49L, 81L, 77L, 70L, 61L, 83L, 48L, 48L, 52L, 56L,
        74L, 47L, 54L, 65L, 45L, 37L, 65L, 77L, 44L, 76L, 52L, 60L, 37L,
        38L, 65L, 81L, 55L, 68L, 49L, 84L, 81L, 72L, 74L, 62L, 75L, 53L,
        58L, 64L, 57L, 61L, 74L, 64L, 46L, 37L, 81L, 83L, 65L, 69L, 54L,
        77L, 73L, 38L, 47L, 50L, 65L, 83L, 45L, 55L, 70L, 49L, 63L, 64L,
        52L, 41L, 36L
    ),
    SITEID = c(
        1L, 1L, 3L, 2L, 1L, 3L, 1L, 1L, 1L, 1L, 3L, 2L, 2L, 1L, 3L, 1L,
        1L, 1L, 1L, 4L, 4L, 3L, 4L, 2L, 1L, 3L, 1L, 3L, 3L, 3L, 3L, 1L,
        1L, 1L, 1L, 3L, 1L, 1L, 2L, 1L, 1L, 1L, 1L, 2L, 1L, 1L, 4L, 1L,
        1L, 3L, 2L, 1L, 4L, 1L, 3L, 2L, 1L, 4L, 1L, 4L, 1L, 3L, 2L, 3L,
        1L, 3L, 4L, 2L, 1L, 2L, 1L, 1L, 3L, 1L, 1L, 4L, 1L, 1L, 1L, 2L,
        1L, 2L, 1L, 3L, 2L, 1L, 1L, 1L, 2L, 1L, 1L, 4L, 3L, 4L, 2L, 1L,
        1L, 1L, 3L, 1L, 1L, 2L, 1L, 4L, 1L, 1L, 4L, 2L, 1L, 2L, 1L, 4L,
        1L, 1L, 3L, 1L, 2L, 1L, 2L, 1L, 1L, 1L, 2L, 1L, 1L, 1L, 1L, 1L,
        1L, 1L, 1L, 2L, 1L, 3L, 3L, 1L, 1L, 1L, 1L, 3L, 1L, 3L, 3L, 4L,
        4L, 1L, 1L, 1L, 1L, 1L, 4L, 1L, 2L, 1L, 4L, 1L, 1L, 4L, 1L, 1L,
        1L, 4L, 4L, 1L, 1L, 1L, 3L, 2L, 1L, 2L, 4L, 4L, 1L, 1L, 1L, 1L,
        3L, 1L, 1L, 3L, 1L, 3L, 3L, 2L, 1L, 2L, 1L, 2L, 3L, 4L, 1L, 3L,
        1L, 3L, 1L, 3L, 1L, 3L, 4L, 3L, 2L, 1L, 2L, 2L, 2L, 1L, 3L, 1L,
        1L, 1L, 3L, 1L, 4L, 1L, 1L, 1L, 1L, 3L, 2L, 1L, 3L, 1L, 2L, 1L,
        3L, 1L, 3L, 1L, 1L, 2L, 1L, 2L, 1L, 2L, 1L, 1L, 1L, 3L, 1L, 1L,
        3L, 1L, 3L, 1L, 4L, 3L, 1L, 1L, 1L, 2L
    ),
    AVAL = c(
        1L, 1L, 0L, 1L, 0L, 1L, 0L, 1L, 0L, 1L, 0L, 1L, 1L, 1L, 0L, 0L,
        1L, 1L, 0L, 0L, 1L, 1L, 0L, 1L, 1L, 0L, 0L, 0L, 0L, 1L, 1L, 0L,
        1L, 1L, 1L, 1L, 0L, 0L, 1L, 0L, 0L, 1L, 1L, 1L, 1L, 0L, 0L, 1L,
        0L, 0L, 1L, 1L, 1L, 0L, 1L, 1L, 1L, 0L, 1L, 0L, 1L, 1L, 1L, 1L,
        0L, 0L, 1L, 1L, 0L, 1L, 1L, 0L, 1L, 1L, 1L, 0L, 1L, 0L, 0L, 1L,
        1L, 0L, 1L, 0L, 1L, 1L, 1L, 1L, 1L, 0L, 1L, 0L, 0L, 0L, 0L, 1L,
        1L, 0L, 1L, 0L, 0L, 0L, 0L, 0L, 0L, 1L, 0L, 1L, 1L, 0L, 0L, 0L,
        1L, 0L, 1L, 0L, 1L, 0L, 1L, 0L, 0L, 0L, 0L, 0L, 0L, 1L, 0L, 1L,
        1L, 1L, 0L, 0L, 0L, 1L, 0L, 0L, 0L, 1L, 0L, 1L, 0L, 1L, 0L, 1L,
        1L, 1L, 1L, 1L, 0L, 1L, 0L, 1L, 1L, 1L, 0L, 1L, 1L, 1L, 0L, 0L,
        1L, 0L, 1L, 0L, 1L, 0L, 0L, 1L, 0L, 0L, 0L, 0L, 0L, 0L, 1L, 1L,
        0L, 0L, 0L, 0L, 1L, 0L, 1L, 1L, 1L, 1L, 1L, 1L, 0L, 0L, 0L, 1L,
        0L, 1L, 1L, 1L, 0L, 0L, 0L, 0L, 1L, 1L, 1L, 0L, 1L, 1L, 1L, 1L,
        0L, 1L, 1L, 0L, 0L, 1L, 0L, 1L, 0L, 0L, 1L, 0L, 0L, 1L, 0L, 0L,
        0L, 1L, 1L, 1L, 0L, 1L, 1L, 1L, 0L, 1L, 0L, 1L, 0L, 0L, 1L, 0L,
        0L, 0L, 0L, 0L, 1L, 0L, 1L, 1L, 0L, 1L
    )
) |>
  mutate(SITEID=factor(SITEID, levels=1L:4L,
                       labels=paste0("Study site ", 1:4)),
         TRT01P=factor(TRT01P, levels=0L:1L, 
                       labels=c("Placebo", "Drug")),
         sAGE = scale(AGE))

trial_data |>
  group_by(SITEID, TRT01P) |>
  summarize(Responders = sum(AVAL),
            n=n(),
            Proportion = Responders/n,
            `Mean age` = mean(AGE),
            `SD age` = sd(AGE),
            .groups="drop") |>
  gt() |>
  fmt_number(columns=c("Mean age", "SD age"), decimals=1) |>
  fmt_number(columns=Proportion, decimals=2)
```

## Model

### Details of approach

We are interested in comparing the two treatment groups in the RCT in terms of
the marginal (population average) difference in the proportion of patients 
achieving the "responder" definition for each treatment.

Obtaining pseudo-random samples from the posterior distribution for a difference
in proportions after fitting a Bayesian logistic regression in `brms` is 
particularly straightforward, if our model only includes treatment. 
In that case, we simply get posterior MCMC samples for the logit for
the treatment and the control group, convert these to probabilities and
form the difference per MCMC sample to obtain MCMC samples for the difference
in proportions. Unlike with some frequentist methods, the properties of 
transformed MCMC samples mean that these samples will automatically obey 
logical constraints (e.g. each proportion must lie in [0, 1], the control 
group proportion plus the difference in proportions must also result in 
a proportion for the treatment group in [0, 1] etc.). This makes it 
extremely simple to obtain a valid point estimate with a sensible 
credible interval.

Things become a little bit more complicated, if there are further terms in 
the model, in the case of our example age and clinical trial site ID. We might
include these terms in our model, because randomization was stratified by 
trial site ("analyze as you randomize") and age is a known important predictor 
of response in this disease.

We now illustrate how we obtain a "standardized" estimator for the
marginal (population average or "unconditional") difference in
proportions using the approach discussed by the [FDA guidance on
adjusting for covariates in randomized clinical trials for drugs and
biological
products](https://www.fda.gov/regulatory-information/search-fda-guidance-documents/adjusting-covariates-randomized-clinical-trials-drugs-and-biological-products)
(see also [@Moore2009] and [@Steingrimsson201718]).  The only thing we do
differently compared to the mentioned papers is that instead of using
a logistic regression model fit with maximum likelihood, we use a
Bayesian logistic regression.

In the Bayesian framework the key steps for obtaining the "standardized" 
estimator of a difference in proportions for a RCT are particularly easy:

1. Fit a model (here a Bayesian logistic regression) to the RCT data.
2. Predict the outcomes for each patient under treatment and control 
(here we obtain pseudo-random samples for the posterior predicted 
logit-probability for each patient).
3. Calculate the metric of interest for each patient for each MCMC sample 
(here we apply the inverse-logit transformation to the logit-probabilities 
and form the difference between the treatment and control ones making sure to
match the same indices so the same underlying set of underlying MCMC samples 
of model parameters are used for the treatment and control prediction of each 
patient).
4. Form the mean across patients.
5. Summarize the resulting MCMC samples for the posterior for the difference
in proportions.


### Model structure

We choose a relatively simple (Bayesian) logistic regression model including
an intercept, treatment, site as a fixed effect, and age as a covariate. We 
use a centered and scaled version of age to make it easier to reason about
priors, because firstly the intercept and site effects now refer to a patient 
with an average age for the study, and secondly the age coefficient now 
describes the change in the logit with every standard deviation (in the study 
population) of age. For full control over the design matrix, we
disable the default automatic centering of the design matrix with the
`center=FALSE` option here.

```{r specifymodel}
#| code-fold: false
#| code-summary: "Our model"
#| output: false
#| warning: false
brms_model1 <- bf(
  AVAL ~ 1 + SITEID + sAGE + TRT01P,
  family = bernoulli(link = "logit"),
  center = FALSE
)
```

<div class="alert alert-info">
  Note that the model used in Step 1 of the procedure for obtaining a 
  "standardized" estimator in the previous subsection could include a
  treatment by baseline covariate interaction. Including such an interaction
  makes sense when (we suspect that) the log-odds-ratio between treatment and 
  control is not constant across all observed baseline covariate values.
</div>

Like for the centered and scaled version of age, we want the intercept term to
not to refer to a reference `SITEID`. Thus, the use sum-to-zero contrasts for 
the `SITEID` factor variable, which means that there is a separate coefficient 
for each factor-level of `SITEID`, but they sum to zero (otherwise the model
would be overparameterized). Sum-to-zero contrasts also mean that we express
a-priori equal uncertainty about the probability for each trial site, while
with a reference category we have less prior uncertainty about the reference
category and more uncertainty about the other categories. With the vague priors
we use in this case study, this consideration may be of minor relevance, but in
other cases it may be more important.

```{r sumtozerocontrast}
#| code-fold: false
#| code-summary: "Setting sum-to-zero contrasts on SITEID factor variable"
contrasts(trial_data$SITEID) <- contr.sum(
  n = length(unique(trial_data$SITEID)),
  contrasts = TRUE
)
```


### Choice of prior distributions

We use a 
[weakly informative prior](https://github.com/stan-dev/stan/wiki/Prior-Choice-Recommendations#prior-for-the-regression-coefficients-in-logistic-regression-non-sparse-case) 
for the regression coefficients.

For the intercept of the model we use a normal distribution that 
(on the logit-scale) approximates a vague Beta(0.5, 0.5) prior distribution for 
a single proportion.

The table below shows what normal and Student-t distributions match the moments
(mean, variance and - for the Student-t distribution - excess kurtosis) 
of the distribution of logit-probabilities for three commonly used vague 
Beta prior distributions for proportions.

```{r visualizeprior}
#| code-fold: true
#| code-summary: "Beta prior approximations"

# Functions to deal with finding appropriate logit-normal SD
# ---------------------
# Density of logit-normal distribution
# (see https://en.wikipedia.org/wiki/Logit-normal_distribution)
dlogitnorm <- function(x, mu = 0, sigma = 1, log = FALSE) {
  tmp_ <- -log(sigma) -
    0.5 * log(2 * pi) -
    (qlogis(x) - mu)^2 / 2 / sigma^2 -
    log(x) -
    log1p(-x)
  if (log) tmp_ else exp(tmp_)
}

# Variance of logit-normal distribution (only for mu=0)
vlogitnorm <- function(mu = 0, sigma = 1) {
  if (mu == 0 & length(sigma) == 1) {
    integrate(
      f = \(x) (plogis(x) - 0.5)^2 * dnorm(x, 0, sigma),
      lower = -Inf,
      upper = Inf
    )$value
  } else if (mu == 0 & length(sigma) > 1) {
    sapply(sigma, FUN = \(y) {
      integrate(
        f = \(x) (plogis(x) - 0.5)^2 * dnorm(x, 0, y),
        lower = 0,
        upper = 1
      )$value
    })
  } else {
    stop("vlogitnorm does not deal with mu neq 0")
  }
}

# Variance of the beta distribution
vbeta <- function(shape1 = 1, shape2 = 1) {
  shape1 * shape2 / (shape1 + shape2)^2 / (shape1 + shape2 + 1)
}
#vbeta(shape1=0.5, shape2=0.5)

# Find SD of the logit-normal to match a particular beta-distribution
find_logitnorm_sd <- function(shape1 = 0.5, shape2 = 0.5) {
  if (all(shape1 != shape2)) {
    stop("Cannot deal with unequal shape parameters")
  } else if (length(shape1) == 1L && length(shape2) == 1L) {
    ucminf::ucminf(par = log(3.14), fn = \(x) {
      (vbeta(shape1 = shape1, shape2 = shape2) - vlogitnorm(sigma = exp(x)))^2
    })$par |>
      exp()
  } else if (length(shape1) == length(shape2)) {
    map2_dbl(shape1, shape2, \(s1, s2) {
      ucminf::ucminf(par = log(3.14), fn = \(x) {
        (vbeta(shape1 = s1, shape2 = s2) - vlogitnorm(sigma = exp(x)))^2
      })$par |>
        exp()
    })
  }
}

tibble(alpha = c(1.0 / 3.0, 0.5, 1.0), beta = c(1.0 / 3.0, 0.5, 1.0)) |>
  mutate(
    `Prior distribution for proportion` = paste0(
      "Beta(",
      round(alpha, 2),
      ", ",
      round(beta, 2),
      ")"
    ),
    scale = sqrt(trigamma(alpha) + trigamma(beta)),
    `Normal(mean, SD) matched on logit-scale` = paste0(
      "N(0, ",
      round(scale, 2),
      ")"
    ),
    scale2 = find_logitnorm_sd(alpha, beta),
    `Normal(mean, SD) matched on probability scale` = paste0(
      "N(0, ",
      round(scale2, 2),
      ")"
    ),
    nu = round(
      4 +
        6 *
          (trigamma(alpha) + trigamma(beta))^2 /
          (psigamma(alpha, 3) + psigamma(beta, 3)),
      1
    ),
    `Student-t(DF, location, scale) matched on logit-scale` = paste0(
      "T(",
      nu,
      ", ",
      0,
      ", ",
      round(scale * sqrt((nu - 2) / nu), 2),
      ")"
    )
  ) |>
  dplyr::select(
    `Prior distribution for proportion`,
    starts_with("Normal"),
    `Student-t(DF, location, scale) matched on logit-scale`
  ) |>
  gt() |>
  tab_spanner(
    label = "Prior distribution for intercept on logit scale (moment matching)",
    columns = c(
      "Normal(mean, SD) matched on logit-scale",
      "Normal(mean, SD) matched on probability scale",
      "Student-t(DF, location, scale) matched on logit-scale"
    )
  )

```

A Beta(1, 1) distribution gives a flat density across all probabilities, a 
Beta(1/2, 1/2) distribution is 
[Jeffreys prior](https://en.wikipedia.org/wiki/Jeffreys_prior) for a single
proportion and Beta(1/3, 1/3) is [@Kerman2011]. It is often
preferable to use at least weakly informative prior distributions instead of
such "default" vague prior distributions. However, note that these distributions
that are frequently quoted as vague prior distributions for a proportion, do
not translate into, say, a Normal(0, 10000) distribution, but rather into 
distributions with scale parameters smaller than 5. 

```{r setprior}
#| code-fold: false
#| code-summary: "Our priors for the model"
#| output: false
#| warning: false
brm_prior1 <- prior(class = b, normal(0, 3.14), coef = Intercept) +
  prior(class = b, student_t(3, 0, 2.5))
```


### Model fitting

```{r fitmodel}
#| code-fold: false
#| code-summary: "Model fitting"
#| output: false
#| warning: false
brmfit1 <- brm(
  formula = brms_model1,
  prior = brm_prior1,
  refresh = 0,
  seed = 3560295,
  data = trial_data
)
```

After fitting this logistic regression model, we obtain the following 
posterior summaries on the logit-scale:

```{r summaryfit}
#| code-fold: false
#| code-summary: "Model fit summary"
summary(brmfit1)
```

Note that the coefficient for fourth site is not shown despite the sum-to-zero 
constraint, but is defined to be `SITEID4 = ( - SITEID1 - SITEID2 - SITEID3 )`.

Let's first pretend our covariate distribution is representative of the target
population of interest. In that case, we predict the causal effect of treatment
for each patient in terms of the difference in proportions for the patient from 
our model (i.e. the difference between the proportion our model predicts for 
each patient on drug and in the control group).

This case also illustrates how using the 
[rvar datatype](https://mc-stan.org/posterior/articles/rvar.html) via the 
`posterior` R package makes our code more concise and easier to understand:

::: panel-tabset

## Using rvars

The `add_epred_rvars` function makes it easy to obtain the expectation 
of the posterior predictive distribution for each patient, including 
our uncertainty around it.

```{r rvarsdiff}
#| code-fold: false
#| code-summary: "Obtain marginal difference in proportions using rvars"
#| cache: false
# Create data with each patient in control and test condition
new_data <- trial_data |>
  mutate(TRT01P = "Placebo") |>
  bind_rows(mutate(trial_data, TRT01P = "Drug")) |>
  mutate(AVAL = 0L)

difference <- new_data |>
  add_epred_rvars(brmfit1, value = "prob") |>
  pivot_wider(
    id_cols = c("USUBJID", "SITEID", "AGE"),
    names_from = "TRT01P",
    values_from = "prob"
  ) |>
  mutate(Difference = Drug - Placebo)

difference

```

## Using predictive rvars

Unlike `add_epred_rvars`, the `add_predicted_rvars` function 
adds draws from the posterior predictive distribution. In this
case (i.e. for a binomial distribution) this results in the 
same mean, but adds the additional variability coming from 
the binomial distribution (as illustrated by the larger
standard deviation of the resulting `rvar`).

```{r rvarsdiffpred}
#| code-fold: false
#| code-summary: "Obtain marginal difference in proportions using rvars of predictives"
#| cache: false
# Create data with each patient in control and test condition
new_data <- trial_data |>
  mutate(TRT01P = "Placebo") |>
  bind_rows(mutate(trial_data, TRT01P = "Drug")) |>
  mutate(AVAL = 0L)

pred_difference <- new_data |>
  add_predicted_rvars(brmfit1, value = "pred") |>
  pivot_wider(
    id_cols = c("USUBJID", "SITEID", "AGE"),
    names_from = "TRT01P",
    values_from = "pred"
  ) |>
  mutate(Difference = Drug - Placebo)

pred_difference

```

## Not using rvars

```{r nonrvarsdiff}
#| code-fold: false
#| code-summary: "Obtain marginal difference in proportions not using rvars"
#| cache: false

# Create data with each patient in control and test condition
new_data <- trial_data |>
  mutate(TRT01P = "Placebo") |>
  bind_rows(mutate(trial_data, TRT01P = "Drug")) |>
  mutate(AVAL = 0L)

# Obtain predicted logit-proportion from model
difference_oldstyle <- posterior_linpred(brmfit1, newdata = new_data) |>
  t() |>
  as_tibble(.name_repair = \(x) paste0("V", 1:length(x))) |>
  bind_cols(new_data |> dplyr::select(USUBJID, TRT01P, SITEID, AGE)) |>
  relocate(USUBJID, TRT01P) |>
  pivot_longer(
    cols = starts_with("V"),
    names_to = "m",
    values_to = "pred_logit"
  ) |>
  mutate(m = as.integer(str_extract(m, "[0-9]+"))) |>
  pivot_wider(
    id_cols = c("USUBJID", "SITEID", "AGE", "m"),
    names_from = "TRT01P",
    values_from = "pred_logit"
  ) |>
  mutate(Difference = inv_logit(Drug) - inv_logit(Placebo))

difference_oldstyle
```

:::

<div class="alert alert-info">
  We ensure that we form the difference in model-predicted proportions (for drug
  minus placebo) for each patient for the same MCMC sample. That way the 
  two  proportions are based on the same MCMC sample from the posterior 
  distribution of the model parameters.
</div>

If we then average this difference across the patients in the study,
we obtain the **marginal (population average) difference in proportions** 
for the study population.

<div class="alert alert-info">
  Note that we obtain the predicted probability on treatment and control for 
  each patient and MCMC sample, form the difference and average these across 
  patients. This is much more sample efficient than the (also valid) alternative 
  of starting by sampling binary outcomes from the posterior predictive 
  distribution on treatment and control for each patient and MCMC sample. 
  The latter approach is inefficient, because we estimate proportions through 
  Monte Carlo integration that we already had. Doing so is valid as we
  aim for the mean proportion difference here. As illustration, the
  predictive approach is also shown resulting in a less precise
  estimate of the population mean difference proportion.
</div>

## Results

### Population average difference in proportions

We can now obtain the **marginal (population average) difference in
proportions** for the study population by summarizing the posterior
samples for the difference in proportions for all patients in the
study population.

::: panel-tabset
## Population average using rvars

```{r rvarspopavg}
#| code-fold: false
#| code-summary: "Summarize marginal difference in proportions via rvars"
#| cache: false
summarize(difference,
          summarize_draws(
            rvar_mean(Difference),
            # Note if we do not explicitly request summary
            # statistics, we would get these, but a 90% credible
            # interval instead of 95%
            "mean",
            "median",
            "sd",
            "mad",
            ~ quantile2(.x, probs = c(0.025, 0.975)),
            "rhat",
            "ess_bulk",
            "ess_tail"
          )) |>
  gt() |>
  fmt_number(
    columns = c("mean", "median", "sd", "mad", "q2.5", "q97.5"),
    decimals = 3
  ) |>
  fmt_number(columns = starts_with("ess"), decimals = 0) |>
  fmt_number(columns = "rhat", decimals = 2)

```

## Population average using predictive rvars

```{r rvarspopavgpred}
#| code-fold: false
#| code-summary: "Summarize marginal difference in proportions via predictive rvars"
#| cache: false
summarize(pred_difference,
          summarize_draws(
            rvar_mean(Difference),
            # Note if we do not explicitly request summary
            # statistics, we would get these, but a 90% credible
            # interval instead of 95%
            "mean",
            "median",
            "sd",
            "mad",
            ~ quantile2(.x, probs = c(0.025, 0.975)),
            "rhat",
            "ess_bulk",
            "ess_tail"
          )) |>
  gt() |>
  fmt_number(
    columns = c("mean", "median", "sd", "mad", "q2.5", "q97.5"),
    decimals = 3
  ) |>
  fmt_number(columns = starts_with("ess"), decimals = 0) |>
  fmt_number(columns = "rhat", decimals = 2)

```

## Population average using rvars (multiple steps)

```{r rvarsmultipopavg}
#| code-fold: false
#| code-summary: "Summarize marginal difference in proportions via rvars"
#| cache: false
# Here we keep two steps of rvars to show how we would first obtain a
# rvar of the average difference, before then getting summaries on it.
summarize(difference, avg_diff = rvar_mean(Difference)) |>
  mutate(summarize_draws(
    avg_diff,
    # Note if we do not explicitly request summary
    # statistics, we would get these, but a 90% credible
    # interval instead of 95%
    "mean",
    "median",
    "sd",
    "mad",
    ~ quantile2(.x, probs = c(0.025, 0.975)),
    "rhat",
    "ess_bulk",
    "ess_tail"
  )) |>
  dplyr::select(-avg_diff) |>
  gt() |>
  fmt_number(
    columns = c("mean", "median", "sd", "mad", "q2.5", "q97.5"),
    decimals = 3
  ) |>
  fmt_number(columns = starts_with("ess"), decimals = 0) |>
  fmt_number(columns = "rhat", decimals = 2)

```

## Population average not using rvars

```{r nonrvarspopavg}
#| code-fold: false
#| code-summary: "Summarize marginal difference in proportions not using rvars"
#| cache: false
difference_oldstyle |>
  dplyr::select(Difference) |>
  posterior_summary() |>
  as_tibble(rownames = "Parameter") |>
  gt() |>
  fmt_number(columns = is.numeric, decimals = 3)
```

:::

### Potential interpretational issues

We can look at whether our model suggests that the difference in proportion 
differs across different values of age and clinical trial site. In this case,
the difference in proportions varies, but the estimate is always somewhat close 
to the overall median of 
`r difference |> summarize(summarize_draws(rvar_mean(Difference))) |> pull(median) |> round(2)`.
The largest deviations are seen for young patients in study site 4 and old
patients in study site 2. It is not really clear whether these differences
are just chance findings due to sampling variability, but plotting the model
predicted proportions for each treatment group separately clarifies this a
little.

::: panel-tabset
### Plot difference by SITEID and age using rvars

```{r rvarsplotdiffsiteid}
#| code-fold: true
#| code-summary: "Summarize marginal difference in proportions via rvars"
#| cache: false
difference |>
  group_by(SITEID, AGE) |>
  summarize(avg_diff = rvar_mean(Difference), .groups = "drop") |>
  mutate(summarize_draws(
    avg_diff,
    ~ quantile2(.x, probs = c(0.5, 0.025, 0.975))
  )) |>
  ggplot(aes(x = AGE, y = q50, ymin = q2.5, ymax = q97.5)) +
  geom_hline(yintercept = 0, lty = 2, color = "darkred") +
  geom_hline(
    yintercept = difference |>
      summarize(avg_diff = rvar_mean(Difference)) |>
      mutate(summarize_draws(avg_diff)) |>
      pull(median),
    lty = 2,
    color = "black"
  ) +
  geom_ribbon(alpha = 0.2, col = "black") +
  geom_line(lwd = 2, col = "black") +
  coord_cartesian(ylim = c(0, 0.35)) +
  facet_wrap(~SITEID) +
  ylab("Posterior model-predicted\ndifference in proportions")

```

### Plot difference by SITEID and age without rvars

```{r nonrvarsplotdiffsiteid}
#| code-fold: true
#| code-summary: "Plot difference in proportions vs model terms"
#| cache: false
difference_oldstyle |>
  group_by(SITEID, AGE) |>
  summarize(
    median = median(Difference),
    lcl = quantile(Difference, probs = 0.025),
    ucl = quantile(Difference, probs = 0.975),
    .groups = "drop"
  ) |>
  ggplot(aes(x = AGE, y = median, ymin = lcl, ymax = ucl)) +
  geom_hline(yintercept = 0, lty = 2, color = "darkred") +
  geom_hline(
    yintercept = median(pull(difference_oldstyle, Difference)),
    lty = 2,
    color = "black"
  ) +
  geom_ribbon(alpha = 0.2, col = "black") +
  geom_line(lwd = 2, col = "black") +
  coord_cartesian(ylim = c(0, 0.35)) +
  facet_wrap(~SITEID) +
  ylab("Posterior model-predicted\ndifference in proportions")
```

:::

As we can see, in the plot below, the largest deviations from the population
average difference in proportion occur when the model predicted proportions 
for one treatment group get close to 0 (younger patients in site 4) or 1 
(older patients in site 2). Overall, the model predicted difference in 
proportions never deviates from the population average by too much, because 
while the predicted proportions for the two treatment groups vary across 
different values of age and clinical trial site, they do not get very close to 
0 or 1 where boundary effects occur.

::: panel-tabset

### Plot group-level predictions by SITEID and age using rvars

```{r rvarsplotdiffsiteidage}
#| code-fold: true
#| code-summary: "Summarize marginal difference in proportions via rvars"
#| cache: false
marginal_difference <- difference |>
  summarize(
    avg_drug = rvar_mean(Drug),
    avg_placebo = rvar_mean(Placebo),
    .by = c(SITEID, AGE)
  )

bind_rows(
  mutate(
    marginal_difference,
    summarize_draws(avg_drug, ~ quantile(.x, probs = c(0.5, 0.025, 0.975)))
  ),
  mutate(
    marginal_difference,
    summarize_draws(avg_placebo, ~ quantile(.x, probs = c(0.5, 0.025, 0.975)))
  )
) |>
  mutate(variable = str_extract(variable, "drug|placebo")) |>
  ggplot(aes(
    x = AGE,
    col = variable,
    fill = variable,
    y = `50%`,
    ymin = `2.5%`,
    ymax = `97.5%`
  )) +
  geom_hline(yintercept = c(0, 1), lty = 2, color = "darkred") +
  geom_ribbon(alpha = 0.3) +
  geom_line() +
  scale_color_brewer(NULL, palette = "Set1") +
  scale_fill_brewer(NULL, palette = "Set1") +
  geom_text(
    data = tibble(
      SITEID = "Study site 1",
      AGE = c(45, 75),
      y = c(0.8, 0.2),
      variable = c("Drug", "Placebo"),
      `2.5%` = NA_real_,
      `97.5%` = NA_real_
    ),
    aes(y = y, label = variable),
    col = RColorBrewer::brewer.pal(n = 3, "Set1")[1:2]
  ) +
  facet_wrap(~SITEID) +
  xlab("Age") +
  ylab("Model predicted proportion") +
  theme(legend.position = "none")
```

### Plot group-level predictions by SITEID and age without rvars

```{r nonrvarsplotdiffsiteidage}
#| code-fold: true
#| code-summary: "Plot predicted proportions vs model terms"
#| cache: false
difference_oldstyle |>
  group_by(SITEID, AGE) |>
  summarize(
    median_pbo = median(inv_logit(Placebo)),
    lcl_pbo = quantile(inv_logit(Placebo), probs = 0.025),
    ucl_pbo = quantile(inv_logit(Placebo), probs = 0.975),
    median_drug = median(inv_logit(Drug)),
    lcl_drug = quantile(inv_logit(Drug), probs = 0.025),
    ucl_drug = quantile(inv_logit(Drug), probs = 0.975),
    .groups = "drop"
  ) |>
  ggplot(aes(x = AGE)) +
  geom_hline(yintercept = c(0, 1), lty = 2, color = "darkred") +
  geom_ribbon(
    fill = rgb(0.2, 0.2, 0.8, 0.3),
    aes(ymin = lcl_pbo, ymax = ucl_pbo)
  ) +
  geom_line(lwd = 2, col = rgb(0.2, 0.2, 0.8, 1), aes(y = median_pbo)) +
  geom_ribbon(
    fill = rgb(0.8, 0.2, 0.2, 0.3),
    aes(ymin = lcl_drug, ymax = ucl_drug)
  ) +
  geom_line(lwd = 2, col = rgb(0.8, 0.2, 0.2, 1), aes(y = median_drug)) +
  geom_text(
    data = tibble(SITEID = "Study site 1", AGE = 45, y = 0.8, label = "Drug"),
    aes(y = y, label = label),
    col = rgb(0.8, 0.2, 0.2, 1)
  ) +
  geom_text(
    data = tibble(
      SITEID = "Study site 1",
      AGE = 75,
      y = 0.2,
      label = "Placebo"
    ),
    aes(y = y, label = label),
    col = rgb(0.2, 0.2, 0.8, 1)
  ) +
  facet_wrap(~SITEID) +
  xlab("Age") +
  ylab("Model predicted proportion")

```

:::

Thus, in this case a marginal (population average) difference in proportion is 
arguably still approximately a
reasonable summary of efficacy for all the studied patients, as well as the 
population. However, it is not clear that this would be the case for 
substantially younger or older patients, or for patients coming from clinical
centers that have higher or lower average response rates than the four study
sites used during the study.

While in the example, the marginal difference in proportions may be a 
reasonable summary measure for efficacy, this would not be the case, if the 
difference varied more substantially across values of predictors for the 
outcome.

<div class="alert alert-info">
  Note that what matters is whether the difference in proportions differs across
  the values of predictors whether they are in the model or not. So, we also
  need to worry about whether we omitted any important predictors when we 
  do this kind of assessment. We will explore this in the next subsection.
</div>

The other interpretational issue is whether the four particular clinical trial
sites adequately and fully represent the distribution of the hospitals or 
doctor's offices, where the drug would be used in clinical practice. 
That is likely not the case.  Firstly, there were just four sites in the RCT
making it hard to believe they could adequately represent all the hospitals 
and physicians' practices that treat the disease under study. Secondly, 
clinical trial sites are usually selected both because they see a sufficient
number of patients with the disease of interest, and because they have the 
necessary experience and infrastructure to participate in a clinical trial.
That suggests that there would likely be a shift in distribution between the 
clinical trial setting and clinical practice. Depending on the size and nature 
of this shift, the non-transportability of the difference in proportions may be
of concern.

### Omitting terms from the analysis model

<div class="alert alert-info"> Keeping the model "simple" by omitting
  terms from the model to avoid too much of a difference in
  model-predicted differences in proportions for patients does not
  address the potential problem, because the marginal difference in
  proportions is still an average across patients (and due to the
  collapsability of the effect measure it is unaffected by including
  or leaving out covariates).
</div>

For comparison purposes, let us fit a model without covariates 
(other than treatment group).

```{r nocovmodel, results='hide',message=FALSE,warning=FALSE}
#| code-fold: false
#| code-summary: "No covariate model"
brms_model2 <- bf(
  AVAL ~ 1 + TRT01P,
  family = bernoulli(link = "logit"),
  center = FALSE
)

brm_prior2 <- prior(class = b, normal(0, 3.14), coef = Intercept) +
  prior(class = b, cauchy(0, 2.5))

brmfit2 <- brm(
  formula = brms_model2,
  prior = brm_prior2,
  data = trial_data,
  seed = 356456,
  refresh = 0
)
```

We can now illustrate that for a model with an intercept and a term
for treatment (but without further covariates) the standardized
estimator of [@Steingrimsson201718] gives identical results to what we
obtain directly from the model parameters via
$\text{logit}^{-1}(\beta_0 + \beta_1) - \text{logit}^{-1}(\beta_0)$,
where $\beta_0$ is the intercept and $\beta_1$ is the coefficient for
active drug.

```{r nocovmodelresults}
#| code-fold: true
#| code-summary: "Different ways of getting marginal proportion difference"
#| cache: false
difference2 <- new_data |>
  add_epred_rvars(brmfit2, value = "prob") |>
  pivot_wider(
    id_cols = c("USUBJID", "SITEID", "AGE"),
    names_from = "TRT01P",
    values_from = "prob"
  ) |>
  mutate(Difference = Drug - Placebo)

# extract posterior as rvar and apply with rdo the inv_logit function
# by draw to the respective rvars. The use of rdo is necessary as
# inv_logit does not accept rvar inputs otherwise. We could also adapt
# the inv_logit function using the rfun adapter. See ?posterior::rfun
difference2a <- with(
  as_draws_rvars(brmfit2),
  rdo(inv_logit(b_Intercept + b_TRT01PDrug) - inv_logit(b_Intercept))
)

summarize_draws(
  c(
    rvar_mean(difference$Difference),
    rvar_mean(difference2$Difference),
    rvar_mean(difference2a)
  ),
  "mean",
  "median",
  "sd",
  "mad",
  ~ quantile2(.x, probs = c(0.025, 0.975)),
  "rhat",
  "ess_bulk",
  "ess_tail"
) |>
  mutate(
    variable = c(
      "with covariates ('standardized' estimator)",
      "without covariates ('standardized' estimator)",
      "without covariates (directly from model parameters)"
    )
  ) |>
  gt() |>
  fmt_number(
    columns = c("mean", "median", "sd", "mad", "q2.5", "q97.5"),
    decimals = 3
  ) |>
  fmt_number(columns = starts_with("ess"), decimals = 0) |>
  fmt_number(columns = "rhat", decimals = 2)

```

As we can see, the marginal treatment difference is very similar (in this case 
slightly attenuated) and
the posterior distribution wider when do no adjust for covariates, as we 
would expect from the literature on the topic such as [@Moore2009] and 
[@Steingrimsson201718]. In the absence of additional covariates, it's 
equivalent to use "standardized" estimator compared with calculating the 
difference in proportions from model parameters, but the "standardized" 
estimator does not require a detailed understanding of the 
parameterization of the model.

## Conclusions

Just as when using maximum-likelihood methods, we can use the "standardized"
estimator for obtaining marginal (population-average) differences in proportions
in the Bayesian setting.
Obtaining credible intervals is easier in the Bayesian setting compared with
the maximum likelihood setting, because we can simply apply transformations to 
MCMC samples.

## Excercise

### Excercise 1

Use the "standardized" estimator to obtain the posterior marginal ratio of
proportions (i.e. $\hat{\pi}_\text{drug} / \hat{\pi}_\text{placebo}$) with a 
credible interval in the example `trial_data`. Explore whether the rate-ratio is 
an appropriate summary measure for the treatment effect in this example.

### Excercise 2

Analyze the data below from a hypothetical two-arm RCT of a blood pressure 
lowering drug compared with placebo with a linear regression model such as
`CHG ~ 1 + TRT01P + BASE`. We might a-priori expect the residual standard 
deviation to be about 12 mmHg. Use this model to obtain the marginal population
average difference in proportions of patients achieving a mean sitting 
systolic blood pressure < 140 mmHg. How does the result compare to fitting
a binomial regression model?

```{r excercisetwo}
#| code-fold: true
#| code-summary: "Data for excercise 2"
bp_trial <- tibble(
    USUBJID = 1L:50L, 
    TRT01P = factor(
      c(2L, 1L, 1L, 1L, 2L, 1L, 1L, 2L, 2L, 2L, 1L, 2L, 2L, 1L, 1L, 2L, 2L, 2L,
        2L, 2L, 1L, 2L, 1L, 1L, 2L, 1L, 2L, 1L, 1L, 2L, 1L, 1L, 1L, 2L, 2L, 2L,
        2L, 2L, 1L, 1L, 1L, 2L, 1L, 1L, 1L, 1L, 1L, 2L, 2L, 2L),
      labels = c("Placebo", "Drug")),
    PARAM = factor("mean sitting Systolic Blood Pressure"),
    BASE = c(
        151.5, 146.6, 142.5, 152.2, 142.9, 140.4, 144.9, 145.5, 142.9, 146, 
        146.8, 148.7, 146.9, 160, 147.8, 145.9, 140.6, 140.6, 145, 141.6, 143.9, 
        143.8, 143.7, 143.9, 141.3, 149.6, 146.6, 149.1, 150.7, 141.9, 142.9, 
        140.5, 155, 145.9, 145, 148.9, 144.9, 149.8, 141.5, 151.8, 146, 140.2, 
        143.6, 140, 140.3, 140.4, 141.2, 145.7, 146.2, 143
    ),
    AVAL = c(
        123.2, 134.8, 123, 140.4, 120.2, 138.7, 153, 142.3, 132.6, 105.9, 137, 
        136.5, 122, 144.1, 135.5, 89.3, 110.2, 140.2, 133.1, 139.6, 126.8, 
        135.6, 136.7, 151.5, 125.6, 133.9, 133.8, 144.1, 145.5, 94, 156.3, 
        136.5, 147.2, 95.9, 127, 146.8, 111.2, 125.6, 125.6, 142.6, 133.7, 
        130.6, 125.8, 138.9, 118.3, 135.1, 123.4, 139.5, 114.7, 138.2
    ),
    CHG = c(
        -28.3, -11.8, -19.5, -11.8, -22.7, -1.7, 8.1, -3.2, -10.3, -40.1, -9.8,
        -12.2, -24.9, -15.9, -12.3, -56.6, -30.4, -0.4, -11.9, -2, -17.1, -8.2, 
        -7, 7.6, -15.7, -15.7, -12.8, -5, -5.2, -47.9, 13.4, -4, -7.8, -50, -18, 
        -2.1, -33.7, -24.2, -15.9, -9.2, -12.3, -9.6, -17.8, -1.1, -22, -5.3, 
        -17.8, -6.2, -31.5, -4.8
    )
)
```
