---
author:
  - Björn Holzhauer - <bjoern.holzhauer@novartis.com>
  - Sebastian Weber - <sebastian.weber@novartis.com>
format:
  html:
    toc: true
    number-sections: true
    html-math-method: katex
bibliography: references.bib
---
  
# Robust borrowing in the meta-analytic combined approach {#sec-rmac}

<!-- https://raw.githubusercontent.com/Novartis/bamdd/main/src/_macros.qmd -->
{{< include _macros.qmd >}}

This case study:
  
- shows how to implement a robust version of the meta-analytic 
combined (MAC) approach 
- discusses how the robust MAC approach for handling prior-data conflict 
by introducing a coefficient for historical studies (with a prior that is 
at the same time spiky near zero and heavy-tailed) compares to the 
robust meta-analytic predictive (rMAP) approach
- introduces continuous scale mixture distributions as robustifying
priors; examples are the double-exponential or Student-$t$
distribution, and shows how to use them both in `brms` and, in the
form of an equivalent normal mixture, in `RBesT`

```{r includestuff}
#| include: false
#| cache: false
here::i_am("src/02n_robust_mac.qmd")
source(here::here("src", "setup.R"))
```

To run the R code of this section please ensure to load these libraries first:
  
```{r basesetup}
#| eval: true
#| echo: true
#| message: false
#| warning: false
#| cache: false
#| code-fold: false
#| code-summary: "Setup code"
library(tidyverse)
library(here)
library(brms)
library(RBesT)
library(gt)
library(ggdist) # visualizing distributions/priors/posteriors
library(distributional) # distributions (that we then use in via ggdist)
library(directlabels)

# instruct brms to use cmdstanr as backend and cache all Stan binaries
options(brms.backend="cmdstanr", cmdstanr_write_stan_file_dir=here("_brms-cache"))
# create cache directory if not yet available
dir.create(here("_brms-cache"), FALSE)
set.seed(678571)

# default controls args passed to Stan
control_args <- list(adapt_delta=0.99, step_size=0.1)

# nice common ggplot2 theme
theme_set(theme_bw(base_size=18))
```

```{r brmscachestuff}
#| include: false
#| cache: false
# speedup things by lowering adapt_delta for now, which speeds up things a lot
# do this only if caching is on (which is when we want things go fast)
# TODO: Can we set adapt_delta to low values based on knowing this is a non-production run??
control_args <- list(adapt_delta=0.8)
```

## Background

When comparing a drug to a control group, we often want to make use of prior
knowledge of what we expect for the control group based on past trials.
We would like methods for this to rely on the historical data when it appears
to be compatible with our new data, but we want to increasingly discard it when
it appears that the past experience may no longer be applicable for the present
situation.

Let us discuss this for a simple case when the behavior of the control group of 
each study can be described with a single parameter. Let us assume that we
are interested in fitting an exponential time to event model in a single new
study and want to borrow information about the control group (log-)hazard rate.

### Meta-analytic predictive (MAP) approach
One popular approach is to fit a hierarchical (meta-analytic) model
to historical studies and to predict control group log-hazard rate for
a new study. The samples from this posterior predictive for the mean
distribution provide us with a meta-analytic predictive (MAP) prior
[@neuenschwander2010].

For an accurate representation of the MAP prior density and the
ability to exactly pre-specify the prior we usually use a mixture
approximation to the MAP prior [@schmidli2014robust] in a subsequent
analyses of our new data. This was illustrated in the "Use of
historical control data" case study (see @sec-use-hist-control-data)
for the case of binomial proportions. Here, for an exponential hazard
rates we might use a mixture of gamma distributions, or for the
control group log-hazard rate a mixture of normal distributions.

### Robust meta-analytic predictive approach

Adding a weakly-informative or vague mixture component to the MAP
prior [@schmidli2014robust] is one approach to making the MAP prior
robust to prior-data-conflict. This robust MAP (rMAP) approach has
become the defacto standard approach in practice, since it mitigates
type I error inflation which can occur whenever the trial data turns
out to be very different wrt to the historical data.

::: {.alert .alert-info}
**"Prior-data conflict"**: A term commonly used to describe a degree 
of incompatibility between an analysis prior distribution and observed 
data. The more unusual the observed data are under the prior 
predictive distribution (or the more the likelihood favors parameter 
values to which the prior assigns relatively little probability), 
the stronger the "prior-data conflict".

In practice, we will often have a limited sample size so that
seemingly relatively extreme data can still occur under ordinary
sampling variability. This can be assessed by comparing the observed
data to the prior predictive distribution of the data implied by the
pior. Thus, it will rarely be possible to conclude that data as
extreme as observed is almost impossible under a given prior
distribution. Thus, "prior–data conflict" is more a conventional
shorthand for a continuum of "prior–data tension" rather than a
situation one can clearly diagnose.

Still, conventional single-component conjugate prior distributions
have bad frequentist operating characteristics when the 
true parameter values are in conflict with the prior. Thus, we
are interested in priors that are "robust" in the sense of 
mitigating such scenarios by increasingly discarding the historical
data the stronger the prior-data conflict is as illustrated with
a simple example below. 

```{r priorllpostplot}
#| echo: false
#| message: false
#| warning: false
#| results: hide
#| fig-show: asis
prior <- mixgamma(
  informative = c(1, 10, 1),
  param = "ab")
rprior <- mixgamma(
  informative = c(0.5, 10, 1),
  vague = c(0.5, 1, 0.1),
  param = "ab")
likelihood1 <- mixgamma(
  informative = c(1, 75, 1),
  param = "ab")
likelihood2 <- mixgamma(
  informative = c(1, 20, 2),
  param = "ab")
posterior <- postmix(priormix=prior, n = 1, m = 75 / 1)
rposterior <- postmix(priormix=rprior, n = 1, m = 75 / 1)
posterior2 <- postmix(priormix=prior, n = 2, m = 20 / 2)
rposterior2 <- postmix(priormix=rprior, n = 2, m = 20 / 2)

bind_rows(
  tibble(Scenario="Clear conflict", `Prior type`="Conventional\nprior",
       x=seq(0, 105, 0.1),
       Prior=dmix(mix=prior, x=x),
       Likelihood=dmix(mix=likelihood1, x=x),
       Posterior=dmix(mix=posterior, x=x)),
  tibble(Scenario="Clear conflict",  `Prior type`="Robust\nprior",
       x=seq(0, 105, 0.1),
       Prior=dmix(mix=rprior, x=x),
       Likelihood=dmix(mix=likelihood1, x=x),
       Posterior=dmix(mix=rposterior, x=x)),
  tibble(Scenario="No apparent conflict", `Prior type`="Conventional\nprior",
       x=seq(0, 30, 0.1),
       Prior=dmix(mix=prior, x=x),
       Likelihood=dmix(mix=likelihood2, x=x),
       Posterior=dmix(mix=posterior2, x=x)),
  tibble(Scenario="No apparent conflict",`Prior type`="Robust\nprior",
       x=seq(0, 30, 0.1),
       Prior=dmix(mix=rprior, x=x),
       Likelihood=dmix(mix=likelihood2, x=x),
       Posterior=dmix(mix=rposterior2, x=x))
 ) %>%
  pivot_longer(Prior:Posterior, names_to="what", values_to="Density") %>%
  mutate(what=factor(what, levels=c("Prior", "Likelihood", "Posterior"))) %>%
  ggplot(aes(x=x, y=Density, color=what, label=what)) +
  theme_bw(base_size=18) +  
  theme(legend.position="bottom", legend.title=element_blank()) +
  geom_line(alpha=0.9) +
  ggthemes::scale_color_colorblind() +
  facet_grid(cols=vars(Scenario), rows=vars(`Prior type`), scales="free_x") +
  xlab("Placebo event rate per 100 patient-years")
```
:::

### Meta-analytic combined (MAC) approach

It is mathematically equivalent to the MAP approach
[@schmidli2014robust] to fit a single hierarchical model to the
historical control group data and the new data at the same time. This
is called the meta-analytic combined (MAC) approach.

In the MAC approach it does not really matter whether our new data consists 
of a single study or multiple studies. I.e. it can be immediately used for the
analysis of a single new study or for a meta-analysis by having a random study
effect on the intercept of a hierarchical exponential time-to-event
model. The MAC approach makes it explicit that we are jointly
modelling the historical data and the trial data of interest. In so
far it has been suggested in the literature @neuenschwander2016 to
generalize the term historical (control) data to co-data referring to
concurrent data. This follows from the joint modelling perspective
which makes it explicit that we model on-trial and external trial data
at the same time removing any distinction between "old" and "new" data.

For the MAP approach this is not as straightforward to create a prior
that can be used for a meta-analysis, even for a single-parameter
case such as exponential time-to-event data. The reason is that, 
if we assume a normally distributed random study effect on the 
intercept of our model on the log-hazard scale, we need to create 
a bivariate mixture approximation to the joint MAP
distribution of the mean and scale parameters of this random effects
distribution as described by @holzhauer2017. This is a non-standard
extension of the MAP approach of @neuenschwander2010, but we could
use e.g. the `RBesT` R package to fit a multivariate normal mixture
to samples from the posterior of the hierachical mean and log-scale
parameters.

### Robust Meta-analytic combined (rMAC) approach

@holzhauer2019 has proposed a robust MAC (rMAC) approach mitigating
prior-data-conflict. This approach outperforms the robust MAP approach
due to its better behavior in cases of moderate prior-data-conflict,
while both rMAP and rMAC perform well when there is no conflict or an
obvious strong prior-data conflict.

Both the rMAP and rMAC are forms of 'dynamic borrowing' in the sense
that how much the old/trial external data influences the new analysis
is dynamically determined by how closely the new data resembles the
old/external data. This can also both be seen as doing model averaging
between a model that assumes exchangability between old/external and
new data, and a model that discards the old data [@holzhauer2019].

## Data

The data used here are the data from @holzhauer2017 and
@holzhauer2019 regarding the cardiovascular safety of
rosiglitazone. The two papers explored methods for how to analyze
major adverse cardiovascular event (MACE) occurrence in a set of 54
randomized controlled trials (RCTs) of rosiglitazone compared with
controls taking into account what we know about control group hazard
rates from the control groups of 64 other trials (for further details
of the data [see @holzhauer2017].

Time to first MACE occurrence in diabetic patients is reasonably well
described by an exponential distribution. Treating the data as
arising from exponentially distributed event times makes it much more
plausible that control group parameters would be exchangeable across
trials of substantially different duration than if we treated the
number of patients with an event as binomially distributed
[@holzhauer2017].


```{r rosidata}
#| code-fold: true
#| code-summary: "Load and summarize data"
rosiglitazone <- read_delim(here("data", "dataset.csv"),
                            delim = ";",
                            show_col_types = FALSE) |>
  # Simple imputation for time to first event or censoring assuming that there 
  # is no overlap in patients with an event and those that drop-out prematurely
  # (other than those that die of an event of interest). A follow-up of half of 
  # the trial duration is assumed for patients with an event and those that 
  # drop-out. We convert the time-scale to patient-years (rather than weeks by 
  # dividing by 52).
  mutate(t = ( tau*(n-ifelse(is.na(z), 0, z)+ifelse(is.na(m), 0, m)-y) + 
                 tau/2*(ifelse(is.na(z), 0, z)-ifelse(is.na(m), 0, m)+y) ) / 52 )

rosiglitazone |>
  mutate(historical=factor(historical, 
                           levels=c(0:1),
                           labels=c("Rosiglitazone studies", 
                                    "Historical studies")),
         test = factor(test, levels=c(1, 0),
                       labels=c("Rosiglitazone", "Control"))) |>
  group_by(historical, test) |>
  summarize(
    Trials = length(unique(trial)),
    Patients = sum(n),
    `Trials with no event in group` = sum(1L*(y==0)),
    `Patients with event` = sum(y),
    `Follow-up (patient-years)` = sum(t),
    `Patients with event per patient-year` = `Patients with event` / `Follow-up (patient-years)`,
    .groups="drop") |>
  gt() |>
  fmt_number(columns=c(Patients, `Follow-up (patient-years)`), decimals = 0) |>
  fmt_number(columns=`Patients with event per patient-year`, decimals = 3)

```


## Model

The model proposed by @holzhauer2019 for the 
log-hazard rate in arm $j \in \{0, 1\}$ of study $i \in I$ is

$$ \log \lambda_{ij} = \beta_0 + \beta_1 \times 1\{ j = 1 \} + \beta_2
\times 1\{ i \in I_\text{historical} \} + \eta_i,$$
$$\eta_i \sim N(0, \sigma^2).$$

It extends the meta-analytic combined approach to borrowing information from 
historical data by robustifying it against prior-data-conflict. The studies 
consist of the set of historical studies $I_\text{historical}$ and the set of 
new studies $I_\text{new}$. $\beta_0$ is the expected log-hazard rate for a 
"new" historical study.
$\beta_1$ is the log-hazard ratio for test treatment ($j=1$) compared with
control ($j=0$). The $\eta_i$ is the usual random study effect on the
intercept modelling the studies exchangably.

The key component for robustifying the model to prior-data conflict is the 
$\beta_2 \times 1\{ i \in I_\text{historical} \}$ term, where $\beta_2$ is a 
coefficient for how historical studies systematically differ from new studies.
We get full borrowing like in the MAC model when $\beta_2$ is near 0, or full 
discarding of the historical data when $\beta_2$ is freely estimated and allowed
to attain very large absolute values. Thus, the prior on $\beta_2$
ciritcally governs the behavior of the model and therfore it's choice
is important and will be explored in detail below.

@holzhauer2019 implemented 
this model using custom Stan code. However, it is straightforward to implement 
this model using `brms` like this:

```{r model1}
#| code-fold: false
#| code-summary: "Model definition"
rmac_model <- bf( y | rate(t) ~ 1 + (1|trial) + test + historical,
                  family=poisson(link="log"), center=FALSE ) 
# This is equivalent to 
# bf( y  ~ 1 + (1|trial) + test + historical + offset(log(t)),
#     family=poisson(link="log"), center=FALSE ) 
# Note: `test` and `historical` are numeric 0/1 indicators
```

::: {.alert .alert-info}
**NOTE**: The likelihood for an exponential
time-to-first-event model for individual patient data is proportional
to the Poisson likelihood for the total number of patients with an
event in each trial arm with a log(sum across patients of time to
first event or censoring)-offset.  These aggregate data for each trial
happen to be the sufficient statistics for exponentially distributed
hazard rates for each trial arm. 
:::
  
### Choice of prior distributions
  
```{r model1prior}
#| code-fold: false
#| code-summary: "Prior specification"
rmac_prior <- prior(class = b, normal(-3.912, 5), coef = Intercept) + # log(0.02) = -3.912
  prior(class = b, coef = test, cauchy(0, 2.5)) +
  prior(class = b, coef = historical, double_exponential(0, 0.5)) +
  prior(class = sd, coef = Intercept, group = trial, normal(0, 1))
```

#### Prior distribution for the intercept

The exponential hazard rates for MACE events have historically been reported
to be approximately 0.02 per patient-year [@diabetes1998],
but we have substantial uncertainty whether this number would be applicable to
the studies in this analysis due to differing inclusion-exclusion criteria and
medical progress since the UKPDS results were published. Thus, we assign a
normal prior centered on $\log(0.02) \approx -3.912$ but with a large standard 
deviation of 5 to allow for much higher or lower log-hazard rates.

#### Prior distribution for the hierarchical scale hyper-parameter

A half-normal $\text{HN}(0, 1)$ prior for the between scale parameter of the 
trial random effect distribution allows for substantial between trial variability 
($\log(10) \approx 2.3$ lies at the `r round((pnorm(log(10), mean=0, sd=1)-0.5)*2, 3)*100` th 
percentile of this prior). It gives only a low probability to very small 
between trial variability ($\log(1.01) \approx 0.00995$ lies at the 
`r round((pnorm(log(1.01), mean=0, sd=1)-0.5)*2, 3)*100` th 
percentile of this prior).

#### Prior distribution for the treatment effect

A wide vague $\text{Cauchy}(0, 2.5)$ prior for the log-hazard ratio between
test and control was used by @holzhauer2019.

#### Prior distribution for how much historical trials differ from new ones ($\beta_2$)

By assigning a prior to $\beta_2$ that is very spiky near 0
(corresponds to full borrowing), but also heavy-tailed (extreme values
far away from 0 result in almost no borrowing from the historical
data), we achieve a behavior that resembles model averaging between
full borrowing or full discarding of the historical data. An extreme
of this idea is the spike-and-slab prior which puts a point mass at 0
with a certain weight and a flat prior for the remaining weight. This
prior has been proposed in the Bayesian variable selection literature
[@Mitchell1988]. However, the spike-and-slab prior is difficult to
sample due to it's discrete nature and the use of an improper flat
prior which has lead to to the developemnt of the regularized
horseshoe prior [@Piironen2017], for example. These more modern priors
are often some form of a normal scale mixture prior (e.g. the
Student-t distribution which is a normal mixture with an inverse-gamma
distributed variance).

The double-exponential (DE) distribution is a normal scale mixture
distribution with an exponentially distributed variance. The density is
one such spiky long-tailed distribution. A $\text{DE}(0, 0.5)$ prior
performed well for an exponential time-to-event setting in the
simulation study of @holzhauer2019.

As we can see in the Figure below, a DE(0, 0.5) has a much spikier peak than a
N(0, 1) distribution that is sometimes used as a prior for regression 
coefficients, but is much longer tailed than a N(0, 0.4) that has a similar
density near zero.

```{r laplacedist}
#| warning: false
#| code-fold: true
#| code-summary: "Code to illustrate Laplace prior"
#| fig-width: 8.5
dlaplace <- function(x, location, scale) {
  1 / (2 * scale) * exp(-abs(x - location) / scale)
}

rlaplace <- function(n, location, scale) {
  tmp_ <- runif(n = n, -0.5, 0.5)
  location - scale * sign(tmp_) * log(1 - 2 * abs(tmp_))
}

ggplot(tibble(x = c(-4, 4)), aes(x)) +
  stat_function(
    fun = purrr::partial(dlaplace, location = 0, scale = 0.5),
    n = 201,
    aes(colour = "DE(0, 0.5)"),
    linewidth = 2,
    alpha = 0.75
  ) +
  stat_function(
    fun = purrr::partial(dnorm, mean = 0, sd = 1),
    n = 201,
    aes(colour = "N(0, 1)"),
    linewidth = 1,
    alpha = 0.75
  ) +
  stat_function(
    fun = purrr::partial(dnorm, mean = 0, sd = 0.4),
    n = 201,
    aes(colour = "N(0, 0.4)"),
    linewidth = 1,
    alpha = 0.75
  ) +
  ylab("Density") +
  scale_colour_discrete("Distribution") +
  guides(colour = guide_legend(position = "inside")) +
  theme(legend.position.inside = c(0.85, 0.79))
```

### Model fitting

We fit the model as follows:

```{r model1fitting}
#| results: hide
#| message: false
#| warning: false
#| code-fold: false
#| code-summary: "Fit model"
rmac_fit <- brm(
  formula = rmac_model,
  data = rosiglitazone,
  prior = rmac_prior,
  seed = 4657578,
  refresh = 0
)
```

### Alternative models

We can also consider a fully stratified model that sets separate vague priors
for the control group in each trial. This model does not borrow any information
between studies in the rosiglitazone meta-analysis or from the historical 
trials, so we omit the historical trials from the data.

```{r model2}
#| results: hide
#| message: false
#| warning: false
#| code-fold: false
#| code-summary: "Define stratified meta-analysis"
stratified_model <- bf(
  y | rate(t) ~ 0 + trial + test,
  family = poisson(link = "log")
)

# we need the names of all coefficients related to the trial factor,
# which we can get by asking brms which priors it will define via the
# get_prior function which returns a data.frame derived object
trial_coefs <- get_prior(stratified_model, filter(rosiglitazone, historical == 0)) |>
  filter(str_starts(coef, "trial")) |>
  pull(coef)

# the prior_string function allows us to set specific priors for a large
# number of factor levels of a categorical variable
stratified_prior <- 
  prior_string("normal(-3.912, 5)", class = "b", coef = trial_coefs) +
  prior(cauchy(0, 2.5), class = b, coef = test)

stratified_fit <- brm(
  formula = stratified_model,
  data = filter(rosiglitazone, historical == 0),
  prior = stratified_prior,
  refresh = 0
)
```

Another option is to discard the historical data, but borrow information across
the control groups of the trials in the rosiglitazone meta-analysis, in that
case we no longer need the `+ historical` term in the model

```{r model3}
#| results: hide
#| message: false
#| warning: false
#| code-fold: false
#| code-summary: "Model with borrowing amongst rosi trials"
nohist_model <- bf(
  y | rate(t) ~ 1 + (1 | trial) + test,
  family = poisson(link = "log"),
  center = FALSE
)

nohist_prior <- prior(class = b, coef = Intercept, normal(-3.912, 5)) +
  prior(class = b, coef = test, cauchy(0, 2.5)) +
  prior(class = sd, coef = Intercept, group = trial, normal(0, 1))

nohist_fit <- brm(
  formula = nohist_model,
  data = filter(rosiglitazone, historical == 0),
  prior = nohist_prior,
  refresh = 0
)
```

Finally, we look at the MAC model without robustification against 
prior-data-conflict.

```{r model4}
#| results: hide
#| message: false
#| warning: false
#| code-fold: false
#| code-summary: "MAC model"
mac_model <- bf(
  y | rate(t) ~ 1 + (1 | trial) + test,
  family = poisson(link = "log"),
  center = FALSE
)

mac_prior <- prior(class = b, coef = Intercept, normal(-3.912, 5)) +
  prior(class = b, coef = test, cauchy(0, 2.5)) +
  prior(class = sd, coef = Intercept, group = trial, normal(0, 1))

mac_fit <- brm(
  formula = mac_model,
  data = rosiglitazone,
  prior = mac_prior,
  refresh = 0
)
```

## Results

### Hazard ratio for rosiglitazone versus control

The posterior distributions for the hazard ratio for rosiglitazone compared
with control from the four different models are shown in the figure
below.

```{r plotresults}
#| code-fold: true
#| code-summary: "Plot posterior distributions for hazard ratio"
#| fig-width: 8.5
model_labels <- c(
  "Stratified model\n(no historical data)",
  "Borrowing information\namong rosiglitazone trials",
  "Meta-Analytic Combined (MAC)",
  "Robust borrowing (rMAC)"
)

tibble(
  fit = list(stratified_fit, nohist_fit, mac_fit, rmac_fit),
  model = factor(
    model_labels,
    levels = rev(model_labels),
    labels = rev(model_labels)
  )
) |>
  rowwise() |>
  mutate(b_test = as_draws_rvars(fit)$b_test) |>
  ggplot(aes(xdist = b_test, y = model)) +
  ggdist::stat_slabinterval() +
  geom_vline(xintercept = 0, color = "darkred", lty = 2) +
  scale_x_continuous(
    breaks = log(c(0.5, 0.75, 1, 1.25, 1.5, 2, 3, 4)),
    labels = c(0.5, 0.75, 1, 1.25, 1.5, 2, 3, 4),
    minor_breaks = NULL
  ) +
  xlab("Hazard ratio for rosiglitazone\ncompared with control") +
  ylab(NULL)
```

While the four analyses do not result in notably different inference, we
can see the robust borrowing (rMAC) approach less heavily relied on the lower
average hazard rate of the historical trials, which resulted in a higher
posterior hazard ratio for the MAC compared with the rMAC approach. 

In fact, the point estimate of the rMAC approach is close to the one we obtain
when borrowing information about control group log-hazard-rates among 
rosiglitazone trials. However, if we do not use historical data, our posterior 
uncertainty is greater - although not by a lot in this case due to the large
amount of data in the rosiglitazone trials.

### Predictive distribution for a new trial

Let's look at the predictive distribution, we get for a new trial using either
the robust MAC approach, the MAP approach or the rMAP approach.

```{r model1afitting}
#| results: hide
#| message: false
#| warning: false
#| code-fold: false
#| code-summary: "Fit model"
# rMAC model fit to historical data only
rmac_hist_fit <- brm(formula = rmac_model,
                     data = filter(rosiglitazone, historical==1),
                     prior = rmac_prior,
                     refresh = 0)
# MAC model fit to historical data only (used to derive MAP / rMAP)
mac_hist_fit <- brm(formula = mac_model,
                    data = filter(rosiglitazone, historical==1),
                    prior = mac_prior,
                    refresh = 0)
```

Note that when the rMAC model is fit to historical data only, 
the intercept $\beta_0$ and the coefficient $\beta_2$ are not 
separately identified by the likelihood. This is intentional. 
Without new studies this adds the prior on $\beta_2$ as 
uncertainty to the predictions for a new study.

#### Robustifying a MAP prior with a scale mixture {#sec-rmac-scale-mixture}

The rMAC approach suggests another, simpler way of making an existing
MAP prior robust. Instead of fitting the rMAC model, we can take the
MAP prior for a single new trial and add the $\text{DE}(0, 0.5)$
"how much do historical trials differ" term to it directly. For a
single new trial this gives exactly the rMAC predictive distribution,
because the $\beta_2$ term simply shifts the control group log-hazard
rate of the new trial by a double-exponentially distributed amount.

Adding a random shift to a random variable means convolving the two
distributions. The obvious way to do this is by simulation: draw
double-exponential random numbers, add them to the MCMC sample from
the MAP distribution and re-fit a mixture to the result. That works,
but it is a poor idea here. The whole point of a heavy-tailed
robustification is what happens *far out in the tails*, and out there
a Monte Carlo sample contains only a handful of draws. The tails of
the resulting prior — the very feature that governs how quickly the
historical data get discarded — would then be determined by sampling
noise and by whatever `automixfit()` happens to fit in a region where
it has almost no information.

We can avoid this entirely and do the convolution analytically. The
construction rests on three facts:

1. **The double-exponential distribution is a normal scale mixture.**
   If the variance $V$ follows an exponential distribution, then a
   normal distribution with that variance is double-exponential,
   $$\text{DE}(0, b) = \int_0^\infty N(0, V) \, p(V) \, \text{d}V,
   \qquad V \sim \text{Exp}\!\left(\tfrac{1}{2b^2}\right),$$
   where the rate is chosen such that $\text{E}[V] = 2b^2$, which is
   the variance of a $\text{DE}(0, b)$ distribution.

2. **The mixing distribution can be discretized deterministically.**
   We split the exponential distribution of $V$ into $N$ strata of
   equal probability $1/N$ and represent each stratum by a single
   variance value, namely the mean of $V$ within that stratum,
   $$V_k = \text{E}[V \mid V \in \text{stratum } k]
   = N \int_{(k-1)/N}^{k/N} Q_V(p) \, \text{d}p,$$
   where $Q_V$ is the quantile function of $V$. The result is an
   $N$-component normal mixture with all components centered on zero.
   Using stratum *means* rather than stratum quantiles is what makes
   this work: it reproduces the variance $2b^2$ of the
   double-exponential distribution **exactly** for any $N$, as
   explained in the box below. Beyond that first moment it is a
   quasi-Monte Carlo (stratified) approximation: it uses no random
   numbers, is exactly reproducible and gets systematically better as
   $N$ grows.

3. **The sum of two independent normal mixtures is a normal mixture.**
   Convolving a mixture with components $(w_i, \mu_i, \sigma_i)$ with
   one with components $(v_j, \nu_j, \tau_j)$ gives all pairwise
   combinations, with weights $w_i v_j$, means $\mu_i + \nu_j$ and
   variances $\sigma_i^2 + \tau_j^2$. `RBesT` already implements this
   internally, so we do not have to code it ourselves.

Putting these together gives a `robustify_scale_mixture()` function
that turns any normal mixture MAP prior into a robustified one,
without any simulation:

::: {.callout-note collapse="true" title="Discrete scale mixture density approximation considerations"}

The mixing distribution enters only through the quantile function
$Q_V$, which we parametrize by the *upper* tail probability
$p = \text{P}(V > v) = 1 - F_V(v)$, with $F_V$ the distribution
function of $V$. Any symmetric, zero-centered normal scale
mixture can therefore be used simply by passing a different
`qvariance(p, scale)` to the code below.

**Why stratum means.** The integral defining $V_k$ is the ordinary
mean integral, written in a different variable. Substituting
$p = 1 - F_V(v)$, so that $\text{d}p = -f_V(v) \, \text{d}v$ and
$v = F_V^{-1}(1 - p) = Q_V(p)$, maps the range of $V$ onto the unit
interval — reversing its orientation, which is why the sign cancels
against the swapped integration limits — and turns
$$\text{E}[V] = \int v \, f_V(v) \, \text{d}v
\quad \text{into} \quad
\int_0^1 Q_V(p) \, \text{d}p,$$
while the same substitution restricted to a slice of $V$ space gives
the $V_k$ above. Since all weights are $1/N$ and
$\sum_{k=1}^{N} \int_{(k-1)/N}^{k/N} = \int_0^1$, the discretization
therefore satisfies
$$\sum_{k=1}^{N} \tfrac{1}{N} V_k = \int_0^1 Q_V(p) \, \text{d}p
= \text{E}[V]$$
*identically*, for any $N$. As the components are centered on zero,
the mixture variance is exactly $\text{E}[V]$, the variance of the
target. Representing each stratum by its midpoint quantile instead
would have no such property and would noticeably understate the
variance.

**The requirement.** Every $V_k$ must be finite. All strata but the
last cover a bounded range of $V$, so only the last one can fail — and
by the identity above that happens precisely when $\text{E}[V]$ does
not exist. Numerically, the integral is taken over the substitution
$p = e^{-t}$, which turns the heavy upper tail into an exponentially
decaying integrand on $(0, \infty)$; integrating over $p$ directly
leaves an integrable singularity at the last stratum boundary that
adaptive quadrature copes with poorly.

**What this means for the usual robustifying priors:**

| prior | mixing variance $V$ | requires |
|:--|:--|:--|
| double-exponential $\text{DE}(0, b)$ | $\text{Exp}(1/(2b^2))$ | always fine |
| Student-$t_\nu$, scale $b$ | $\text{Inv-Gamma}(\nu/2, \nu b^2/2)$ | $\nu > 2$ |
| horseshoe-like: half-$t_\nu$ on the *scale* | $b^2 F(1, \nu)$ | $\nu > 2$ |
| horseshoe | $b^2 F(1, 1)$ | never |

The last two rows are the same construction, $N(0, b^2\lambda^2)$ with
$\lambda \sim \text{half-}t_\nu(0,1)$, so that $\lambda^2$ is $F$
distributed and $\nu$ tunes how heavy the tails are. The horseshoe is
its $\nu = 1$ (half-Cauchy) special case, which has no mean and is
thus out of reach — and note that it fails *silently*, since
`tan(pi/2)` is a large finite number in floating point rather than
`Inf`. In code:

```r
qvariance_student_t <- function(p, scale, nu = 4) {
  1 / qgamma(p, shape = nu / 2, rate = nu * scale^2 / 2)
}
qvariance_quasi_horseshoe <- function(p, scale, nu = 3) {
  scale^2 * qf(p, df1 = 1, df2 = nu, lower.tail = FALSE)
}
```

**If $\text{E}[V]$ does not exist**, a different discretization is
needed, one that approximates the variance integral stably — for
example representing each stratum by its median
$Q_V\!\left(1 - (k - \tfrac{1}{2})/N\right)$, which is always finite
but no longer reproduces the variance exactly. Finally, the heavier
the tails, the larger $N$ has to be before the far tail of the
approximation settles down.

:::

```{r robustifyscalemixture}
#| code-fold: true
#| code-summary: "Definition of robustify_scale_mixture()"


# Quantile function of the mixing variance V for a DE(0, scale)
# distribution, parametrized by the UPPER tail probability
# p = P(V > v). A DE(0, b) is a N(0, V) mixed over
# V ~ Exp(rate = 1 / (2 * b^2)), where the rate is chosen such that
# E[V] = 2 * b^2, the variance of a DE(0, b).
qvariance_laplace <- function(p, scale) {
  qexp(p, rate = 1 / (2 * scale^2), lower.tail = FALSE)
}

# Discrete normal scale mixture approximation to a symmetric,
# zero-centered normal scale mixture distribution.
#
# We split the distribution of the mixing variance V into n strata of
# equal probability 1/n and represent each stratum by the conditional
# mean of V within it, which is n times the integral of the quantile
# function of V over the stratum. See the box above for why stratum
# means, why E[V] must exist and why we substitute p = exp(-t).
mixnorm_scale_mixture <- function(scale, n = 50,
                                  qvariance = qvariance_laplace) {
  # stratum boundaries as t = -log(p); t_0 = 0 and t_n = Inf
  t_k <- c(0, -log1p(-seq_len(n) / n))
  integrand <- function(t) {
    p <- exp(-t)
    out <- qvariance(p, scale = scale) * p
    # p * Q(p) tends to 0 whenever E[V] exists; this guards against
    # 0 * Inf = NaN once p underflows
    out[!is.finite(out)] <- 0
    out
  }
  v_k <- vapply(seq_len(n), \(k) {
    n * integrate(integrand, t_k[k], t_k[k + 1])$value
  }, numeric(1))
  do.call(mixnorm, lapply(seq_len(n), \(k) c(1 / n, 0, sqrt(v_k[k]))))
}

# Robustify a normal mixture prior by convolving it with a symmetric,
# zero-centered normal scale mixture (by default a double-exponential
# distribution with the given scale).
#
# mix        normal mixture (e.g. as returned by automixfit) to robustify
# scale      scale of the robustifying distribution
# n          number of components used to approximate the mixing
#            distribution
# qvariance  function(p, scale) giving the quantile function of the
#            mixing variance as a function of the upper tail
#            probability p = P(V > v); swap this out for other normal
#            scale mixtures
robustify_scale_mixture <- function(mix, scale, n = 50,
                                    qvariance = qvariance_laplace) {
  # the normal scale mixture trick only works on an unconstrained
  # (normal) scale, so we insist on a normal mixture here
  stopifnot(inherits(mix, "normMix"))
  robust_part <- mixnorm_scale_mixture(scale, n, qvariance)
  # RBesT:::mixnormdiff is an internal RBesT function returning the
  # normal mixture of the *difference* of two normal mixtures; negating
  # the means turns it into the sum we want (a no-op for a symmetric
  # zero-centered mixing distribution, but kept for generality)
  robust_part[2, ] <- -robust_part[2, ]
  RBesT:::mixnormdiff(mix, robust_part)
}
```

How good is the approximation? The figure below compares the exact
$\text{DE}(0, 0.5)$ density with the discrete normal scale mixture for
a few choices of $N$. Note the log-scale for the density, which lets
us judge the tails.

```{r laplaceapprox}
#| code-fold: true
#| code-summary: "Code to check the accuracy of the approximation"
#| warning: false
#| fig-width: 8.5
laplace_check <- tibble(N = c(5L, 15L, 50L)) |>
  rowwise() |>
  reframe(
    N = N,
    x = seq(0, 6, 0.02),
    Approximation = dmix(mixnorm_scale_mixture(scale = 0.5, n = N), x)
  ) |>
  mutate(N = factor(N))

ggplot(laplace_check, aes(x = x)) +
  stat_function(
    fun = purrr::partial(dlaplace, location = 0, scale = 0.5),
    n = 201, aes(linetype = "Exact DE(0, 0.5)"), linewidth = 1.5, alpha = 0.4
  ) +
  geom_line(aes(y = Approximation, colour = N), linewidth = 0.8) +
  scale_y_log10() +
  scale_colour_discrete("Number of\ncomponents N") +
  scale_linetype_manual(NULL, values = "solid") +
  ylab("Density")
```

Even a handful of components does well in the body of the
distribution, and $N = 50$ (the default) is essentially
indistinguishable from the exact density over the range that matters
in practice. Two things are worth pointing out. First, the
approximation slightly rounds off the sharp peak at zero, since a
finite mixture of normals cannot reproduce the cusp of the
double-exponential density. This is irrelevant here, because we
immediately convolve with a MAP prior whose own components have a
non-negligible standard deviation, which smooths the peak away in any
case. Second, very far out the approximation becomes light-tailed
relative to the exact density — increasing $N$ pushes this point
further out.

Since the result is an ordinary `RBesT` normal mixture, we can use it
with all the usual machinery, in particular conjugate updating with
`postmix()`. The price we pay is that the mixture has
$N \times$ (number of MAP components) components, but that causes no
practical difficulty.


We can now fit mixture approximations to these predictive
distributions. We will also construct a robustified version of the MAP
prior directly, by convolving it with a $\text{DE}(0, 0.5)$
distribution as described in @sec-rmac-scale-mixture. This is another
way of achieving the same effect as the rMAC prior.

```{r getrmap}
#| code-fold: false
#| code-summary: "Create different predictive distributions"

map_approx_mac_sample <- posterior_linpred(
  mac_hist_fit,
  newdata = tibble(trial = "newrosi", test = 0, historical = 0, t = 1),
  allow_new_levels = TRUE,
  sample_new_levels = "gaussian"
)

map_approx_mac <- automixfit(map_approx_mac_sample[, 1], Nc = seq(1, 10, 1))

# robustify the MAP prior by convolving it with a DE(0, 0.5) distribution
map_approx_plus_de <- robustify_scale_mixture(map_approx_mac, scale = 0.5)

map_approx_rmap <- robustify(map_approx_mac, weight = 0.3, mean = -3.97, sigma = 5)

map_approx_rmac_sample <- posterior_linpred(
  rmac_hist_fit,
  newdata = tibble(trial = "newrosi", test = 0, historical = 0, t = 1),
  allow_new_levels = TRUE,
  sample_new_levels = "gaussian"
)

map_approx_rmac <- automixfit(map_approx_rmac_sample[,1], Nc = seq(1, 10, 1))

```

If we plot these different predictive distributions, we can see that, as 
expected, both the rMAP and rMAC priors are substantially more heavy-tailed than 
the MAP prior. This is what makes them more robust than the MAP prior to 
prior-data-conflict. However, the two are more long-tailed in a different 
manner: the rMAC prior has a lot of distributional weight on moderate deviations from
the MAP prior, while the rMAP prior puts a lot more prior weight on more extreme
deviations. This has the consequence that the transition from not discounting
the historical data much to completely discounting is quite abrupt for the rMAP
approach, while the transition between the two behaviors is more gradual for
the rMAC prior.

We can also see, that convolving the MAP prior with a
double-exponential distribution reproduces the predictive
distribution of the rMAC approach for a single trial.

```{r plotpreddist}
#| code-fold: true
#| code-summary: "Plot different predictive distributions"
#| fig-width: 8.5
#| fig-height: 6
xbreaks <- c(0.0001, 0.001, 0.01, 0.1, 0.5)
ggplot(tibble(logHR = log(c(1E-4, 1))), aes(logHR)) +
  stat_function(
    fun = purrr::partial(dmix, mix = map_approx_mac),
    n = 201,
    aes(colour = "MAP"),
    linewidth = 1,
    alpha = 0.75
  ) +
  stat_function(
    fun = purrr::partial(dmix, mix = map_approx_rmap),
    n = 201,
    aes(colour = "rMAP"),
    linewidth = 1,
    alpha = 0.75
  ) +
  stat_function(
    fun = purrr::partial(dmix, mix = map_approx_rmac),
    n = 201,
    aes(colour = "rMAC"),
    linewidth = 1,
    alpha = 0.75
  ) +
  stat_function(
    fun = purrr::partial(dmix, mix = map_approx_plus_de),
    n = 201,
    aes(colour = "MAP \u2295 DE(0, 0.5)"),
    linewidth = 1,
    alpha = 0.75
  ) +
  scale_colour_discrete("Prior") +
  scale_x_continuous(
    breaks = round(log(xbreaks), 1),
    sec.axis = sec_axis(
      ~ exp(.),
      breaks = xbreaks,
      labels = xbreaks,
      name = "hazard-rate"
    )
  ) +
  xlab("log hazard-rate") +
  ylab("Density") +
  labs(title = "Prediction for placebo in new trial of rosiglitazone") +
  theme(legend.position = "bottom")
```

## Conclusion


Methods for dynamic borrowing from external data for both the analysis
of a single trial and for meta-analyses are a topic of ongoing
research. However, `brms` and `RBesT` let us easily implment a number
of these approaches.

In the rosiglitazone example the inference barely changed across the
different approaches models. However, that's partially because there
are a lot of data in the rosiglitazone trials included in the
meta-analysis. The value of dynamic borrowing of historical
information will be greatest when there are a lot of historical data
and not so much data in the current trial(s).

The robust MAP approach is popular due to its simplicity, the
possibility to analyze new data without MCMC sampling and the
intuitively appealing idea of having a tunable weight for a vague
mixture component. However, the weights given to the different mixture
components do not have a simple interpretation. Additionally, the rMAP
approach may not behave well in cases where it is not completely clear
whether there is a prior-data conflict or not.

The robust MAC approach is an interesting approach, because it is both
easy to use and simulation studies suggest that it outperforms the
rMAP approach for borrowing information about exponential
control-group log-hazard rates from historical data. In the recurrent
events setting or for binary outcomes, the same prior
$\text{DE}(0,0.5)$ prior that performed well for exponential
time-to-event data might plausibly also be good choice. In contrast,
there is not yet experience with the rMAC approach for continuous
data. The reader is invited to explore this topic on their own in the
second excercise for this chapter.

Inspired by the rMAC approach, we have another way of robustifying a
MAP prior for a single trial. We can convolve the MAP mixture with a
double-exponential distribution. By exploiting that the
double-exponential distribution is a normal scale mixture, and by
discretizing the mixing distribution deterministically using a
quasi-Monte Carlo approach, this convolution can be done analytically
and yields another normal mixture.  This avoids the Monte Carlo error
that a simulation-based convolution would introduce precisely in the
tails, which are the part of the prior that governs how quickly the
historical data are discarded. It also avoids having to run MCMC at
all, once the MAP prior is available, which is particularly convenient
when evaluating operating characteristics at the design stage.

Taking either samples from the posterior predictive distribution for a
single new trial from the rMAC approach or after robustfying a MAP
prior in that way (which is preferred as it more accuratley reflects
the important heavy-tails of the DE), we can still use conjugate
updating for mixture approximations to these samples when analyzing a
single new trial. While the ability to do so (rather than use MCMC
sampling) may not matter much for the final analysis of a trial, it is
very useful when evaluating operating characteristics at the trial
design stage.  In contrast, when we construct a hyper-prior for a
meta-analysis we would use MCMC sampling whether we use a MAP, rMAP,
MAC or rMAC approach.


## Exercises

1. @beaten2013 reported a RCT of the anti-interleukin-17A monoclonal
antibody secukinumab in the treatment of ankylosing spondylitis. The
endpoint under analysis is a responder endpoint (20\% response
according to the Assessment of SpondyloArthritis international Society
criteria aka "ASAS20" responders at week 6). Use the historical data
available in the `RBesT::AS` data as additional information for the
control group via the rMAC approach.  Obtain samples from the prior
predictive distribution for the control group of the 
@beaten2013 trial and visualize them.
Look at the prior predictive distribution for the treatment and
control groups, as well as the difference in proportions. Analyze the
new trial and compare the results to the posterior by-group
proportions (0.592 vs. 0.245) and difference in proportions (0.347
with 95\% CrI 0.115 to 0.564) reported by @beaten2013.

```{r example2}
example2 <- RBesT::AS |>
  mutate(historical = 1, test = 0) |>
  bind_rows(tibble(
    study = "Baeten et al. 2013",
    r = c(14, 1),
    n = c(23, 6),
    historical = 0,
    test = c(1, 0)
  ))

```

2. The data from @banbeta2022 is about comparing resveratrol to
control in the ADC-37 Alzheimer's disease trial in terms of change
in the continuous Alzheimer's Disease Assessment Scale–Cognitive
(ADAS-Cog) score at week 52. 
Note that a `DE(0, 0.5)` prior on the log-hazard (or log-rate / 
logit scales) is dimensionless, but here has units on the change 
or mean scale. Thus, consider how in the rMAC approach the
width of a double-exponential prior on the control group mean change
from baseline should depend e.g. on the residual standard deviation 
of the endpoint.

```{r adcsexample}
adcs <- tibble(
  trial = c(
    "ADC-11",
    "ADC-15",
    "ADC-16",
    "ADC-22",
    "ADC-27",
    "ADC-37",
    "ADC-37"
  ),
  historical = c(1, 1, 1, 1, 1, 0, 0),
  test = c(0, 0, 0, 0, 0, 0, 1),
  n = c(111, 202, 169, 63, 164, 55, 64),
  chg = c(8.7, 8.0, 4.4, 6.1, 5.1, 4.8, 2.9),
  sd = c(7.2, 5.8, 6.4, 7.1, 6.8, 6.3, 9.6)
)
```

